---
title: "Module 12: Inference"
author: 
  - name: "Adam Spiegler"
    affiliation: "University of Colorado Denver"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    toc-expand: 1
    embed-resources: true
editor: source
#execute:
#  eval: false
title-block-banner: true
warning: false
license: "CC BY-NC-SA"
---


## Additional Reading


- [Chapter 6 Linear Model Inference and Prediction](https://jfrench.github.io/LinearRegression/inference.html#overview-of-inference-and-prediction) from *A Progressive Introduction to Linear Models*
- Chapter 6 of *Linear Regression Models: Applications in R* ([link to library ebook](https://go.exlibris.link/Q7wgCRrz))
- Chapter 6 of *Applied Linear Regression* ([link to library ebook](https://go.exlibris.link/69jsQwWW))


# Loading Required Packages and Data


```{r}
#| message: false
library(tidyverse)
library(faraway)  # access gala data frame
```


```{r}
data(gala)
head(gala)
```

```{r}
summary(gala)
```

The data set `gala` contains 30 observations (one for each island in the Galapagos). There are 6 variables in the data set. The relationship between the number of plant species and several geographic variables is of interest.

- `Species`: The number of plant species found on the island.
- `Area`: The area of the island ($\mbox{km}^2$).
- `Elevation`: The highest elevation of the island (m).
- `Nearest`: The distance from the nearest island (km).
- `Scruz`: The distance from Santa Cruz Island (km).
- `Adjacent`: The area of the adjacent island ($\mbox{km}^2$).


```{r}
lmod <- lm(Species ~ Area + Elevation + Nearest +
            Scruz + Adjacent, data = gala)
summary(lmod)
```

## Question 1  {#sec-q1}


Write out a formula for the fitted model based on the output above.

### Solution to Question 1


$$\widehat{\mbox{Species}} = ?? + ?? \mbox{Area} + ?? \mbox{Elevation} + ?? \mbox{Nearest} + ?? \mbox{Scruz} + ?? \mbox{Adjacent}$$

<br>  


## Question 2 {#sec-q2}

Which regressor has the strongest association with the number species on the island?

### Solution to Question 2


<br>  



# Setting Up The Hypotheses

How can we decide whether all or some of the regressor variables should be included in our model?


- Let $\Omega$ denote the proposed full model.
- Let $\omega$ denote a simpler model that uses a subset of all regressors.

## Question 3 {#sec-q3}

What do you think would be suitable hypothesis to test whether $\Omega$ is a better model than $\omega$?

### Solution to Question 3


<br>  


# Measuring Significance

- If the models have a similar fit, then we prefer $\omega$ since it is simpler.
- If the model $\Omega$  is a much better model than $\omega$, then we prefer $\Omega$.
- If $\mbox{RSS}_{\omega} - \mbox{RSS}_{\Omega}$ is large, then $\Omega$ has a superior fit.
- How large of a difference is large enough?

$$\mbox{test stat} = \frac{\mbox{fit of observed model} - \mbox{fit of null model}}{\mbox{variability due to sampling}} = \frac{\mbox{RSS}_{\omega} - \mbox{RSS}_{\Omega}}{RSS_{\Omega}}.$$


# General $F$ Test for Comparing Two Nested Regression Models:


## Set Up Hypotheses

H~0~: Model $\omega$ is an adequate model.

H~a~:  Model $\Omega$ is better.

## Compute the Test Statistic

Suppose that model $\Omega$ has $p$ estimated regression coefficients and model $\omega$ has $q$ estimated regression coefficients. Then we use rescaling to get an
[**F-statistic**]{style="color:blue"} which has [**F-distribution**]{style="color:blue"} under the null hypothesis given by:

$$F =  \frac{(\mbox{RSS}_{\omega} - \mbox{RSS}_{\Omega})/(df_{\omega}-df_{\Omega})}{RSS_{\Omega}/df_{\Omega}} =  \frac{(\mbox{RSS}_{\omega} - \mbox{RSS}_{\Omega})/(df_{\omega}-df_{\Omega})}{\hat{\sigma}_{\Omega}^2} \sim F_{p-q,n-p} ,$$
where $df_{\omega} = n -q$,  $df_{\Omega} = n -p$, and $\hat{\sigma}_{\Omega}^2 = RSS_{\Omega}/df_{\Omega}$.


Are any of the predictors useful? We compare the full model $\Omega$ given by $\mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\epsilon}$ to the null model $\omega$ which is the constant mean model $y = \mu + \epsilon$ which we estimate with $\hat{y} = \bar{y}$.


- Set up the hypotheses:

  - $H_0$: The null model $\omega$ is adequate.  $\beta_1 = \beta_2 = \ldots = \beta_{p-1} = 0$.
  - $H_a$: At least one of the regression coefficients is not zero.

- Calculate the $F$-statistic setting $q=1$.

$$F =  \frac{(\mbox{RSS}_{\omega} - \mbox{RSS}_{\Omega})/(df_{\omega}-df_{\Omega})}{RSS_{\Omega}/df_{\Omega}} =  \frac{(\mbox{TSS} - \mbox{RSS})/(p-1)}{RSS/(n-p)} = \frac{\mbox{SS}_{\rm reg}/(p-1)}{RSS/(n-p)}.$$

- Calculate the $p$-value using $F_{p-q,n-p}$ where $\mbox{df}_1 = p-q$ is the number of additional regressors in $\Omega$ and $\mbox{df}_2=n-p=\mbox{df}_{\Omega}$ (setting $q=1$). 


# Analysis of Variation (ANOVA) Table

The information above is often presented in an [**analysis of variation (ANOVA) table**]{style="color:blue"} which is convenient for summarizing and organizing the various ingredients in the $F$-statistic. Below is a typical ANOVA table when $q=1$.

| Source | Deg. of Freedom | Sum of Squares | Mean Square | F |
|--------|-----------------|----------------|-------------|----|
| Regression | $p-1$ | $\mbox{SS}_{\rm reg}$ | $\mbox{SS}_{\rm reg}/(p-1)$ | $F$ |
| Residual | $n-p$ | RSS | $\mbox{RSS}/(n-p)$  | |
| Total | $n-1$ | TSS | | |

- **Source**: The source of the variation in the data.
- **Regression**: The variability due to the variable of interest (which is `Area` in this example). Sometimes, the variable is a factor and this row is labeled **Treatment** or **Between**. 
- **Residuals**: The unexplained random error (of the full model $\Omega$). Sometimes when the variable of interest is a factor, the row heading is labeled as **Within** to make it clear that the row concerns the variation within the groups.
- **Total**: The total variation in the data from the null (constant mean) model $\omega$.


## Comparing with Fit of Null Model

```{r}
nullmod <- lm(Species ~ 1, data = gala) 
anova(nullmod, lmod)
```

## Question 4 {#sec-q4}

What is the first line of code above doing?

### Solution to Question 4



<br>  

## Question 5 {#sec-q5}

Interpret the output from the `anova` function.

### Solution to Question 5


<br>  

## Question 6 {#sec-q6}

Arrange the output of the anova function below into an ANOVA table such as the one shown earlier.


### Solution to Question 6


```{r}
(ss <- 381081-89231) # SS_{reg}
(mean.ss <- ss/5) #regression mean sum squares
(mean.rss <- 89231/24) #residual mean sum squares
(fstat <- mean.ss/mean.rss)
```


Replace each ?? with an appropriate value based on output above.


| Source | Deg. of Freedom | Sum of Squares | Mean Square | F |
|--------|-----------------|----------------|-------------|---|
| Variability due to `Area` | ?? | ?? | ?? | ?? |
| Error of $\Omega$ | ?? | ?? | ?? | |
| Error of $\omega$ | ?? | ?? | | |


## Question 7  {#sec-q7}

Compare these calculations with the `summary` output from earlier.

### Solution to Question 7


<br>  


## Verifying the Output Again

```{r}
# Compute the sum of squares
(rss0 <- deviance(nullmod)) # RSS of null (TSS)
(rss <- deviance(lmod)) # RSS of proposed model

# Compute the degrees of freedom
(df0 <- df.residual(nullmod)) # df of null model
(df <- df.residual(lmod)) #  df of proposed model

# Compute the F-statistic
(f <- ((rss0 - rss)/(df0 - df))/(rss/df)) # F-stat

# Compute the P-value with pf
# pf(q, df1, df2) computes area to left of q
# This is a right-tail test (so 1-pf)
1 - pf(f, df1 = df0 - df, df2 = df) # P-value
```

## Compare with Summary of `lm`

```{r}
summary(lmod)
```


## Visualizing the F-dist

```{r}
#| echo: false
#| message: false
#| warning: false
#library(tidyverse)

data.frame(f = 0:1000 / 100) %>% 
           mutate(df_25_26 = df(x = f, df1 = 25, df2 = 26),
                  df_05_26 = df(x = f, df1 = 5, df2 = 26),
                  df_05_10 = df(x = f, df1 = 5, df2 = 10)) %>%
#                  df_05_10 = df(x = f, df1 = 5, df2 = 10)) %>%
  gather(key = "df", value = "density", -f) %>%

  ggplot() +
  geom_line(aes(x = f, y = density, color = df, linetype=df)) +
  labs(title = "F at Various Degrees of Freedom",
       x = "F",
       y = "Density") + 
  xlim(0, 5) + 
  theme_bw()
```

- If $\mbox{df}_1 = p-1$ is constant and we increase $\mbox{df}_2 = \mbox{df}_{\Omega} = n-p$, the center hump is centered around the same $F$ value, but there is more area in the hump and less in the right tail.
  - Same number of predictors but more observations has less variability!

- If $\mbox{df}_2 = n-p$ is constant and we increase $\mbox{df}_1 = \mbox{df}_{\Omega} = p-1$, the center hump shifts to the right and there is more area concentrated near the peak and less in the tail to the right.


## Interpreting the Results

## Question 8 {#sec-q8}

You perform a hypothesis test to test the following claims:

H~0~: The null model $\omega$ is adequate.  $\beta_1 = \beta_2 = \ldots = \beta_{p-1} = 0$.

H~a~: At least one of the regression coefficients is not zero.

In the Galapagos example, the $F$-statistic is $15.699$ and the $p$-value is given as $6.838e-07\ast\ast\ast$


### Question 8a 

Interpret the meaning of the $p$-value. 

#### Solution to Question 8a


<br>  


### Question 8b

If you fail to reject $H_0$, does this mean there is no relation between the regressors and the response?

#### Solution to Question 8b



<br>  


### Question 8c

If you reject $H_0$, does this mean you have found the best model for $y$? 

#### Solution to Question 8c



<br>  


# Testing a Single Regressor


To test whether a single regressor (regressor j) can be dropped from the model, we choose between $H_0:\beta_j = 0$ and $H_a: \beta_j \ne 0$.

We have two options in this case:

- Use the previous approach, letting the reduced model be the one without that regressor.
- Use a $t$-statistic approach.


## Question 9 {#sec-q9}

Repeat the previous process to test whether the regressor `Area` is significant (assuming all others are in the model).

- State the hypotheses.
- Compute the $F$-statistic and $p$-value.
- State the conclusion in this context.


### Solution to Question 9


<br>  


## Using a t-statistic

An alternative approach is to use a $t$-statistic for testing the hypotheses:

$$ t_i = \frac{\widehat{\beta}_i-0}{\mbox{se}(\widehat{\beta}_i)}= \frac{\widehat{\beta}_i}{\mbox{se}(\widehat{\beta}_i)}$$
using a $t$-distribution with $\mbox{df}_{\Omega}=n-p$ degrees of freedom.

```{r}
# calculation of t-stat
beta.area <- summary(lmod)$coeff[2,1]
se.area <- summary(lmod)$coeff[2,2]
(t <- beta.area/se.area)

# verifying p-value
2*pt(t, df.residual(lmod))

# matching F-stat
t^2
```

## Question 10 {#sec-q10}

What is the difference between the test being run in the code below and the previous test? Does the test below lead to the same conclusion?

### Solution to Question 10


```{r}
# compare results to test of beta_Area = 0 when no other predictors in model
summary(lm(Species ~ Area, gala))
```


<br>  



# Testing a Pair of Regressors

## Question 11 {#sec-q11}

Test whether the `Area` and `Adjacent` regressor variables should be simultaneously dropped from the model that already includes `Elevation`, `Nearest`, and `Scruz` in the model.  Make sure to specify the regressors that are the model when stating H~0~ and H~a~.

- State the hypotheses. 
- Compute the $F$-statistic and $p$-value.
- State the conclusion in this context.


### Solution to Question 11


<br>  

