---
title: "Module 14: Collinearity"
author: 
  - name: "Adam Spiegler"
    affiliation: "University of Colorado Denver"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    toc-expand: 1
    embed-resources: true
editor: source
#execute:
#  eval: false
title-block-banner: true
warning: false
license: "CC BY-NC-SA"
---


## Additional Reading


- Chapter 10 of *Linear Regression Models: Applications in R* ([link to library ebook](https://go.exlibris.link/Q7wgCRrz))

# Loading Required Packages and Data


```{r}
#| message: false
library(tidyverse)
library(car)  # compareCoefs and vif functions
library(mctest)  # eigprop function
```

# Motivating Example: Driver Seat Settings

Car drivers like to adjust the seat position for their own comfort. Car designers would find it helpful to know where different drivers will position the seat depending on their size and age. Researchers at the HuMoSim laboratory at the University of Michigan collected data on 38 drivers. They measured `Age` in years, `Weight` in pounds, height with shoes and without shoes on cm (`HtShoes` and `Ht`), seated height `Arm` length (in cm), `Thigh` length (in cm), lower `Leg` length (in cm), and `hipcenter` which is the horizontal distance of the midpoint of the hips from a fixed location in the car in mm.

::: {.callout-caution}
You do not need to load `faraway` with a `library()` command to load the `seatpos` data frame in the code cell below as long as you have previously installed the `faraway` package. However, if you have not installed the `faraway` package, the code will crash. In which case, run the command `install.packages("faraway")` in the console below. Then run the code cell below.
:::


```{r}
data(seatpos, package="faraway")
summary(seatpos)
```


Below we regress `hipcenter` on all other variables in `seatpos`.

```{r}
# fit model with all predictors
lmod1 <- lm(hipcenter ~ ., data = seatpos)
summary(lmod1) 
```

```{r}
# create model matrix which includes the intercept
x <- model.matrix(lmod1)
```

## Question 1 {#sec-q1}

Does $R^2$ value of the model seem consistent with the p-values for each coefficient? Why or why not?

### Solution to Question 1


<br>  
<br>  

## Question 2 {#sec-q2}

What are some methods we have use to check for correlations between regressors?


### Solution to Question 2


<br>  
<br>  


## Question 3 {#sec-q3}

Do you see any large pairwise correlations? Do these make practical sense in this context?

### Solution to Question 3


<br>  
<br>  


# Collinearity

Collinearity concerns covariability among the regressors.  When the columns of the model matrix $X$ are linearly dependent, then $X^T X$ is singular and there is no unique least squares estimate of $\beta$.

- The columns of $X$ are said to be **exactly collinear** in this case.
- This causes serious problems with estimation and interpretation.

Even when the columns of $X$ are not perfectly dependent, we may still have problems.

## Why is Collinearity Problematic?

Measuring `hipcenter` is difficult. Suppose the measurement error had a standard deviation of 10 mm. Let's see what happens if we add a little bit of measurement error to the response.

```{r}
set.seed(1) # allows us to reproduce results

# add a little measurement error
lmod1a <- lm(hipcenter + 10 * rnorm(38) ~ ., data = seatpos) 
summary(lmod1a)

#library(car)  # compareCoefs is from car package
compareCoefs(lmod1, lmod1a, se = FALSE) # compare coefficients
```


## Question 4 {#sec-q4}

Comment on the changes to the coefficients from Model 1 and Model 2.

### Solution to Question 4

<br>  
<br>  



# Methods for Detecting Collinearity

We have too many variables! We can reduce the collinearity by removing carefully chosen variables in an appropriate order to get a more stable model, and more clear interpretation of the remaining predictors. We should not conclude that the variable(s) we remove from the model are not related to the response.


## Examine the Correlation Matrix

Examine the pairwise correlation matrix of the regressors and look for *large* pairwise correlations.

- Large is a bit subjective, but the larger the correlation among regressors, the more likely it is that you have a collinearity problem.

```{r}
round(cor(seatpos[,-9]), 3)
```

If we want to include `Weight` in our model, then we could say any other regressor with a correlation coefficient $| r | > 0.7$ will be removed from the model. Then we could remove `HtShoes`, `Ht`, `Seated`, and `Leg`.

With multiple linear regression models we need to move beyond bivariate analysis comparing pairs of regressors. To assess multicollinearity, we need to consider whether one regressor is a near linear combinations of a collection of other regressors.

## Regress on Predictors


Let $R_j^2$ denote the $R^2$ value from an auxiliary regression model with $X_j$ as the response variable and all the other variables (excluding the response) as the regressors. The quantity $1-R_j^2$ is called the [tolerance]{style="color:blue"}. 


## Question 5 {#sec-q5}

Run the two code cells below and identify the values of $R_{\rm \texttt{HtShoes}}^2$ and $R_{\rm \texttt{Age}}^2$. How can we use these values to determine whether or not we should include `HtShoes` and/or `Age` in the model for `hipcenter`?

```{r}
summary(lm(HtShoes ~ Age + Weight + Ht + Seated + Arm + Thigh + Leg, data = seatpos))
```

```{r}
summary(lm(Age  ~ Weight + HtShoes + Ht + Seated + Arm + Thigh + Leg, data = seatpos))
```

### Solution to Question 5

<br>  
<br>  


## The Variance Inflation Factor (VIF)

The effect of collinearity is that the some regression coefficients have a large variance. The variance of $\widehat{\beta}_j$ can be expressed in the form

$$\begin{align}
\mbox{Var}(\widehat{\beta}_j ) &= \frac{ \sum (y_i - \hat{y}_i)^2}{\sum (x_i - \bar{x})^2 (1 - R_j^2)(n-k-1)}\\
&= {\color{blue}{\frac{1}{1-R_j^2}}} \frac{ \sum (y_i - \hat{y}_i)^2}{\sum (x_i - \bar{x})^2 (n-k-1)} \\
&= ({\color{blue}{\mbox{VIF}_j}}) \ \frac{ \sum (y_i - \hat{y}_i)^2}{\sum (x_i - \bar{x})^2 (n-k-1)} .
\end{align}$$



The [**variance inflation factor (VIF)**]{style="color:blue"} of regressor $j$ is 

$$\mbox{VIF}_j = \frac{1}{1-R_j^2}.$$

The larger the correlation of $X_j$ with the other regressors:

- The closer $R_j^2$ is to 1, and thus
- The larger the variance inflation factor, VIF. 

**The VIF is the standard diagnostic for assessing collinearity.**

- A $\mbox{VIF}_j$ **more than 10 indicates a potential problem with collinearity** for regressor $X_j$.
- Other more conservative values for the VIF, such as $5$, can be commonly used as well.
- Orthogonality with other predictors implies that $R_j^2=0$ and $\mbox{VIF}_j = 1$, which minimizes the variance of $\widehat{\beta}_j$.



## Question 6 {#sec-q6}

Using the output from [Question 5](#sec-q5), compute the VIF for `HtShoes` and `Age`.

### Solution to Question 6


<br>  
<br>  


### Compute VIF with `vif()`

Using the `vif()` function in the `car` package, we can compute the VIF's of all regressors.


## Question 7 {#sec-q7} 

Run the code cell below and then decide which predictors seem to have a problem with collinearity?

```{r}
#library(car)
vif(lmod1)  # calculate the vifs
```

### Solution to Question 7


<br>  
<br>  



### Interpreting VIF

When interpreting the VIF, it is more practical to interpret the value of $\sqrt{\mbox{VIF}}$ similar to how we typically take the square root of the variance and interpret the standard deviation. $\sqrt{\mbox{VIF}}$ represents the increased width of the 95% CI for the slope $\widehat{\beta}_j$. For instance:

- $\mbox{VIF}_{\rm HtShoes} = 307.4$ suggests that the 95% CI
for $\hat{\beta}_{\rm HtShoes}$ is $\sqrt{307.4}=17.5$ times as wide as it would be if the regressors had a correlation of zero (no collinearity).


## Question 8 {#sec-q8} 

Play around with model and see what happens when your remove correlated variables.

```{r}
vif(lmod1)
```

### Solution to Question 8


<br>  
<br>  



## Examine the Eigenvalues of $X^TX$.


Eigenvalues provide a measure of joint variability among a set of explanatory variables. Eigenvalues of $X^TX$ close to $0$ indicate high multicollinearity because they correspond to near-linear dependencies among the predictor variables. An eigenvalue of $0$ would denote perfect multicollinearity and a singular matrix $X$.

Let $\lambda_1 \leq \lambda_2 \leq \ldots \leq \lambda_p$ be the eigenvalues of the regressors (including the intercept) ordered from largest to smallest.

- For observational data, when there is a **wide range in the eigenvalues**, this indicates there is a **potential problem with collinearity**.
- When the [**condition number**]{style="color:blue"} $\kappa= \sqrt{\frac{\lambda_1}{\lambda_p}} \geq 30$ then **there is a potential problem with collinearity**.


```{r}
# model matrix off all observed regressors values
# except for the intercept column we ignore for
x <- model.matrix(lmod1)[,-1]
lam <- eigen(t(x) %*% x)  #  eigenvalues of X^TX
round(lam$val, 3) 
```


```{r}
# compute condition number
sqrt(lam$val[1] / lam$val[8])
```

A [**condition index**]{style="color:blue"} $\kappa_j= \sqrt{\frac{\lambda_1}{\lambda_j}}$ shows the degree of multicollinearity in a regression design matrix. 

```{r}
# compute condition indices
(kappa <- round(sqrt(lam$val[1]/lam$val), 3))
```



### Using SVD

[**Singular values**]{style="color:blue"} of $X$ are the square roots of the eigenvalues of $X^TX$. 


```{r}
svd_result <- svd(x) # SVD of model matrix
round(svd_result$d^2, 3)  # square of singular values
```



## Scaling Regressors

Belsley[^1] recommends that when using condition indices to assess collinearity that:

- The intercept be included in your $X$ matrix.
- The columns of $X$ should NOT be centered.
- The columns of $X$ should be scaled (i.e., the standard deviation of each column should be constant).

[^1]: Belsley, D.A. Computer Science in Economics and Management (1991)


```{r}
# scale all variables except hipcenter
seatpos[, -9] <- scale(seatpos[, -9], center = FALSE, scale = TRUE)
head(seatpos)
```


```{r}
lmod_sc <- lm(hipcenter ~ ., data = seatpos)  # model with scaled regressors
x_sc <- model.matrix(lmod_sc)  # model matrix includes intercept column
lam_sc <- eigen(t(x_sc) %*% x_sc)  # eigenvalues of X^TX
round(lam_sc$val, 6)  # eigenvalues of scaled model
``` 


```{r}
# condition indices for scaled model
(kappa_sc <- round(sqrt(lam_sc$val[1]/lam_sc$val), 3))
```


- The other condition indices $\sqrt{\frac{\lambda_1}{\lambda_j}}$ are worth examining because they may indicate a problem with more than one linear combination of the regressors.
- This does not tell us which regressors may be correlated, only that collinearity is a problem.

### Which Regressors are Leading to Large Condition Indices?

[**Variance decomposition proportions**]{style="color:blue} (or variance proportion, VP) can be examined to determine the regressors that are leading to large condition indices. Larger variance decomposition proportions (VP's) suggest a greater proportion of a variable’s variance is accounted for by other regressors.To determine whether sufficient multicollinearity is present to warrant concern:

- Look for any condition index greater than or equal to $30$.
- If one is identified, look across its row and find any VP that exceeds $0.50$. 
A condition index greater than or equal to $30$ coupled with a VP greater than or equal to $0.50$ indicates a collinearity or multicollinearity issue that should be analyzed further. 

::: {.callout-note}
If you have not installed the `mctest` package, in the console run the command `install.packages("mctest")`. Then run the command `library(mctest)` to load the package. 

- Run `?eigprop` from the console to access help documentation.
:::


```{r}
#library(mctest)  # make sure mctest has been installed and loaded
eigprop(lmod1, Inter = TRUE, prop = 0.5) #both are default options.
```


## Question 9 {#sec-q9}

Based condition indices and the variance decomposition proportions in the output above, which regressor should be first? Then continue removing variables until you believe collinearity has been addressed.


### Solution to Question 9


<br>  
<br>  



**You should assess collinearity right after exploratory data analysis and before variable selection.**

- If your regressors are collinear, then all the subsequent inference is suspect and none of the diagnostics require you to fit a model first.
- It is better to remove collinear variables first, then proceed with analysis.  

## Question 10 {#sec-q10}

What methods can be utilized to look for and fix problems with collinearity?


### Solution to Question 10


<br>  
<br>  




