---
title: "Solutions 12: Inference"
author: 
  - name: "Adam Spiegler"
    affiliation: "University of Colorado Denver"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    toc-expand: 1
    embed-resources: true
editor: source
#execute:
#  eval: false
title-block-banner: true
warning: false
license: "CC BY-NC-SA"
---


## Additional Reading


- [Chapter 6 Linear Model Inference and Prediction](https://jfrench.github.io/LinearRegression/inference.html#overview-of-inference-and-prediction) from *A Progressive Introduction to Linear Models*
- Chapter 6 of *Linear Regression Models: Applications in R* ([link to library ebook](https://go.exlibris.link/Q7wgCRrz))
- Chapter 6 of *Applied Linear Regression* ([link to library ebook](https://go.exlibris.link/69jsQwWW))


# Loading Required Packages and Data


```{r}
#| message: false
library(tidyverse)
library(ellipse)
library(faraway)  # access gala data frame
```


::: {.callout-note}
We have not used the `ellipse` package yet this semester. If you have not installed this package, be sure to run the command `install.packages("ellipse")` from the console below and re-run the code cell above.
:::

```{r}
data(gala)
head(gala)
```

```{r}
summary(gala)
```

The data set `gala` contains 30 observations (one for each island in the Galapagos). There are 6 variables in the data set. The relationship between the number of plant species and several geographic variables is of interest.

- `Species`: The number of plant species found on the island.
- `Area`: The area of the island ($\mbox{km}^2$).
- `Elevation`: The highest elevation of the island (m).
- `Nearest`: The distance from the nearest island (km).
- `Scruz`: The distance from Santa Cruz Island (km).
- `Adjacent`: The area of the adjacent island ($\mbox{km}^2$).

# Hypothesis Testing for a Simple Linear Regression


## Question 1  {#sec-q1}

Write out a formula for the fitted model based on code and output below.

```{r}
lmod_area <- lm(Species ~ Area, data = gala)
summary(lmod_area)
```

## Setting Up The Hypotheses


## Question 2 {#sec-q2}

What do you think would be suitable hypothesis to test whether the predictor `Area` should be included?

### Solution to Question 2

- H~0~: There is no linear association between the `Species` and `Area`. $\beta_1 =0$.
- H~a~: There is a linear association between the `Species` and `Area`. $\beta_1 \ne 0$.


## Calculating the Test Statistic

If the null hypotheses is true, how likely would we be to obtain the observed sample data?

$$t = \frac{\text{observed} - \text{null}}{\text{SE}} = \frac{\widehat{\beta}_1 - 0}{\text{SE}(\widehat{\beta}_1)}.$$ 

Recall the OLS estimate for $\hat{\beta}_1$ is

$$\hat{\beta}_1 = \frac{\sum (x_i - \bar{x})(y - \bar{y})}{\sum (x_i - \bar{x})^2}.$$


In regression, the standard error can be approximated by the formula:

$$
\text{SE}(\widehat{\beta}_1) = \sqrt{ \frac{\sum (\hat{y}_i - y_i)^2 / (n-2)}{\sum (x_i - \bar{x})^2 }}.
$$


- $\frac{\sum (\hat{y}_i - y_i)^2$ is the RSS of the model with `Area` as the only predictor.
- $n-2$ denotes the degrees of freedom with $2$ being the number of regression coefficients.
- $\sum (x_i - \bar{x})^2$ is the total sum of the squared deviance in the predictor.

## Question 3 {#sec-q3}

Approximate the standard error for $\widehat{\beta}_1$. Then use this result to compute the test statistic and p-value corresponding to this test.


### Solution to Question 3

```{r}
# standard error
n <- length(gala$Species)
x_bar <- mean(gala$Area)
top <- deviance(lmod_area) / (n-2)
bottom <- sum((gala$Area - x_bar)^2)
sqrt(top / bottom)
```


Thus, we have test statistic

$$t = \frac{\widehat{\beta}_1 - 0}{\text{SE}(\widehat{\beta}_1)} = \frac{0.08196 - 0}{0.01971} \approx  4.1583$$


```{r}
# compute test statistic
0.08196 / 0.01971
```


```{r}
# compute P-value
2 * (1 - pt(4.158295, n - 2))
```


```{r}
summary(lmod_area)
```

Since the p-value $0.00027$ is less than the significance level chosen (5%), the test is statistically significant. We can reject $H_0$ and accept $H_a$. In this context we can say the linear association between `Area` and `Species` is significant, and the slope coefficient corresponding to `Area`. is not equal to zero.

If we square the $t$ test statistic, we get 

```{r}
(0.08196 / 0.01971)^2
```

which is the value of the $F$-statistic stated in the summary output of the linear model.


# Hypothesis Tests with Multiple Linear Regression

## Question 4  {#sec-q4}

Write out a formula for the fitted model based on code and output below.

```{r}
lmod <- lm(Species ~ Area + Elevation + Nearest +
            Scruz + Adjacent, data = gala)
summary(lmod)
```


### Solution to Question 4



$$\widehat{\mbox{Species}} = 7.068 -  0.0239 \mbox{Area} + 0.3195 \mbox{Elevation} + 0.0091 \mbox{Nearest} -0.2405 \mbox{Scruz} - 0.0748 \mbox{Adjacent}$$


## Question 5 {#sec-q5}

Which regressor has the strongest association with the number species on the island? Explain how you determined your answer.

### Solution to Question 5

- Hard to tell at this point. We can see the `Elevation` had the largest slope (in absolute terms) and the smallest p-value, so that seems important!

- The regressor `Scruz` has a large slope (in absolute terms), however the variability is quite large as indicated by the standard error, and thus we see it has a relatively large p-value, indicating that is plausible the coefficient is actually 0 or it could be positive.



# Setting Up The Hypotheses

How can we decide whether all or some of the regressor variables should be included in our model?


- Let $\Omega$ denote the proposed **complete or full model**.
- Let $\omega$ denote a **simpler model** that uses a subset of all regressors.

## Question 6 {#sec-q6}

What do you think would be suitable hypothesis to test whether $\Omega$ is a better model than $\omega$?

### Solution to Question 6

- H~0~: Simple Model $\omega$ is adequate.
- H~a~: Complete Model $\Omega$ is preferred.


# Measuring Significance

- If the models have a similar fit, then we prefer $\omega$ since it is simpler.
- If the model $\Omega$  is a much better model than $\omega$, then we prefer $\Omega$.
- If $\mbox{RSS}_{\omega} - \mbox{RSS}_{\Omega}$ is large, then $\Omega$ has a superior fit.
- How large of a difference is large enough?

## An Initial Test Statistic

$$\mbox{test stat} = \frac{\mbox{fit of observed model} - \mbox{fit of null model}}{\mbox{variability due to sampling}} = \frac{\mbox{RSS}_{\omega} - \mbox{RSS}_{\Omega}}{RSS_{\Omega}}.$$


# General $F$ Test for Comparing Two Nested Regression Models:


## Set Up Hypotheses

H~0~: Model $\omega$ is an adequate model.

H~a~:  Model $\Omega$ is better.

## Compute the Test Statistic

Suppose that model $\Omega$ has $p$ estimated regression coefficients and model $\omega$ has $q$ estimated regression coefficients. Then we use rescaling to get an
[**F-statistic**]{style="color:blue"} which has [**F-distribution**]{style="color:blue"} under the null hypothesis given by:

$$F =  \frac{(\mbox{RSS}_{\omega} - \mbox{RSS}_{\Omega})/(df_{\omega}-df_{\Omega})}{RSS_{\Omega}/df_{\Omega}} =  \frac{(\mbox{RSS}_{\omega} - \mbox{RSS}_{\Omega})/(df_{\omega}-df_{\Omega})}{\hat{\sigma}_{\Omega}^2} \sim F_{p-q,n-p} ,$$
where $df_{\omega} = n -q$,  $df_{\Omega} = n - p$, and $\hat{\sigma}_{\Omega}^2 = RSS_{\Omega}/df_{\Omega}$.


::: {.callout-note}
We divide each $RSS$ by its degrees of freedom to calculate an **unbiased** estimate of the variance similar to how we divide by $n-1$ when finding an unbiased estimate for the standard deviation. This prevents underestimation.
:::

Are any of the predictors useful? We compare the full model $\Omega$ given by $\mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\epsilon}$ to the null model $\omega$ which is the constant mean model $y = \mu + \epsilon$ which we estimate with $\hat{y} = \bar{y}$.


- Set up the hypotheses:

  - $H_0$: The null model $\omega$ is adequate.  $\beta_1 = \beta_2 = \ldots = \beta_{p-1} = 0$.
  - $H_a$: At least one of the regression coefficients is not zero.

- Calculate the $F$-statistic setting $q=1$.

$$F =  \frac{(\mbox{RSS}_{\omega} - \mbox{RSS}_{\Omega})/(df_{\omega}-df_{\Omega})}{RSS_{\Omega}/df_{\Omega}} =  \frac{(\mbox{TSS} - \mbox{RSS})/(p-1)}{RSS/(n-p)} = \frac{\mbox{SS}_{\rm reg}/(p-1)}{RSS/(n-p)}.$$

- Calculate the $p$-value using $F_{p-q,n-p}$ where $\mbox{df}_1 = p-q$ is the number of additional regressors in $\Omega$ and $\mbox{df}_2=n-p=\mbox{df}_{\Omega}$ (setting $q=1$). 


```{r}
#| echo: false
#| eval: false
# Use pf() for areas under F dist
1 - pf(15.7, 5, 24)
```


# Analysis of Variation (ANOVA) Table

The information above is often presented in an [**analysis of variation (ANOVA) table**]{style="color:blue"} which is convenient for summarizing and organizing the various ingredients in the $F$-statistic. Below is a typical ANOVA table when $q=1$.

| Source | Deg. of Freedom | Sum of Squares | Mean Square | F |
|--------|-----------------|----------------|-------------|----|
| Regression | $p-1$ | $\mbox{SS}_{\rm reg}$ | $\mbox{SS}_{\rm reg}/(p-1)$ | $F$ |
| Residual | $n-p$ | RSS | $\mbox{RSS}/(n-p)$  | |
| Total | $n-1$ | TSS | | |

- **Source**: The source of the variation in the data.
- **Regression**: The variability due to the variable of interest (which is `Area` in this example). Sometimes, the variable is a factor and this row is labeled **Treatment** or **Between**. 
- **Residuals**: The unexplained random error (of the full model $\Omega$). Sometimes when the variable of interest is a factor, the row heading is labeled as **Within** to make it clear that the row concerns the variation within the groups.
- **Total**: The total variation in the data from the null (constant mean) model $\omega$.


## Comparing with Fit of Null Model

```{r}
nullmod <- lm(Species ~ 1, data = gala) 
anova(nullmod, lmod)
```

## Question 7 {#sec-q7}

What is the first line of code above doing?

### Solution to Question 7

It is creating the null model which is our model $\omega$ if we are testing all regressors. This is the constant mean model $y = \beta_0 + \epsilon = \mu + \epsilon$.

## Question 8 {#sec-q8}

Interpret the output from the `anova` function.

### Solution to Question 8

We see the $F$-statistic is $15.699$ which has a $p$-value $=6.838 \times 10^{-7}$ which is extremely small. We see that our test is statistically significant, and we have evidence to support the alternative hypothesis that at least one of the regression coefficients is not zero, though we cannot state specifically which regressor coefficient(s) are likely to be nonzero.

## Question 9 {#sec-q9}

Arrange the output of the anova function below into an ANOVA table such as the one shown earlier.


### Solution to Question 9


```{r}
(ss <- 381081-89231) # SS_{reg}
(mean.ss <- ss/5) #regression mean sum squares
(mean.rss <- 89231/24) #residual mean sum squares
(fstat <- mean.ss/mean.rss)
```



Replace each ?? with an appropriate value based on output above.


| Source | Deg. of Freedom | Sum of Squares | Mean Square | F | 
|--------|-----------------|----------------|-------------|---|
| Variability due to `Area` | $5$ | $291850$ | $58370$ | $15.69948$ |
| Error of $\Omega$ | $24$ | $89231$ | $3717.958$ | |
| Error of $\omega$ | $29$ | $381081$ | | |


## Question 10  {#sec-q10}

Compare these calculations with the `summary` output from earlier.

### Solution to Question 10


They match!


## Verifying the Output Again

```{r}
# Compute the sum of squares
(rss0 <- deviance(nullmod)) # RSS of null (TSS)
(rss <- deviance(lmod)) # RSS of proposed model

# Compute the degrees of freedom
(df0 <- df.residual(nullmod)) # df of null model
(df <- df.residual(lmod)) #  df of proposed model

# Compute the F-statistic
(f <- ((rss0 - rss)/(df0 - df))/(rss/df)) # F-stat

# Compute the P-value with pf
# pf(q, df1, df2) computes area to left of q
# This is a right-tail test (so 1-pf)
1 - pf(f, df1 = df0 - df, df2 = df) # P-value
```

## Compare with Summary of `lm`

```{r}
summary(lmod)
```


## Visualizing the F-dist

```{r}
#| echo: false
#| message: false
#| warning: false
#library(tidyverse)

data.frame(f = 0:1000 / 100) %>% 
           mutate(df_25_26 = df(x = f, df1 = 25, df2 = 26),
                  df_05_26 = df(x = f, df1 = 5, df2 = 26),
                  df_05_10 = df(x = f, df1 = 5, df2 = 10)) %>%
#                  df_05_10 = df(x = f, df1 = 5, df2 = 10)) %>%
  gather(key = "df", value = "density", -f) %>%

  ggplot() +
  geom_line(aes(x = f, y = density, color = df, linetype=df)) +
  labs(title = "F at Various Degrees of Freedom",
       x = "F",
       y = "Density") + 
  xlim(0, 5) + 
  theme_bw()
```

- If $\mbox{df}_1 = p-1$ is constant and we increase $\mbox{df}_2 = \mbox{df}_{\Omega} = n-p$, the center hump is centered around the same $F$ value, but there is more area in the hump and less in the right tail.
  - Same number of predictors but more observations has less variability!

- If $\mbox{df}_2 = n-p$ is constant and we increase $\mbox{df}_1 = \mbox{df}_{\Omega} = p-1$, the center hump shifts to the right and there is more area concentrated near the peak and less in the tail to the right.


## Interpreting the Results

## Question 11 {#sec-q11}

You perform a hypothesis test to test the following claims:

H~0~: The null model $\omega$ is adequate.  $\beta_1 = \beta_2 = \ldots = \beta_{p-1} = 0$.

H~a~: At least one of the regression coefficients is not zero.

In the Galapagos example, the $F$-statistic is $15.699$ and the $p$-value is given as $6.838e-07\ast\ast\ast$


### Question 11a 

Interpret the meaning of the $p$-value. 

#### Solution to Question 11a


There is an extremely slim chance (essentially $0\%$) that would would observe so much of the variability in the response explained by model $\Omega$ compared to $\omega$ if $\Omega$ was just as good of a model as $\omega$.


### Question 11b

If you fail to reject $H_0$, does this mean there is no relation between the regressors and the response?

#### Solution to Question 11b

No, it means the relation is not likely to be linear, but there certainly could another (nonlinear) relation besides a linear relation between the predictors and response.**

Not significant may also imply there is not enough data to confidently conclude a regressor helps describe the mean response.

### Question 11c

If you reject $H_0$, does this mean you have found the best model for $y$? 

#### Solution to Question 11c

Even if we conclude H~a~, we’re not sure that model $\Omega$ is the best model, it is simply preferable to model $\omega$.

- Not all regressors may be necessary.
- Additional regressors may improve the model further.

The F test for a regression relationship is just the beginning of analysis!


# Testing a Single Regressor


To test whether a single regressor (regressor j) can be dropped from the model, we choose between $H_0:\beta_j = 0$ and $H_a: \beta_j \ne 0$.

We have two options in this case:

- Use the previous approach, letting the reduced model be the one without that regressor.
- Use a $t$-statistic approach.


## Question 12 {#sec-q12}

Repeat the previous process to test whether the regressor `Area` is significant (assuming all others are in the model).

- State the hypotheses.
- Compute the $F$-statistic and $p$-value.
- State the conclusion in this context.


### Solution to Question 12

#### Hypotheses

- $H_0: \beta_1=\beta_{\rm Area}=0 \mid \beta_0, \beta_2, \beta_3, \beta_4, \beta_5 \in \mathbb{R}$

- $H_a: \beta_1 \neq 0\mid \beta_0, \beta_2, \beta_3, \beta_4, \beta_5 \in \mathbb{R}$

#### Compute the $F$-statistic and $p$-value.

```{r}
# fit reduced model (full without Area)
lmods <- lm(Species ~ Elevation + Nearest + Scruz + Adjacent, data = gala)
anova(lmods, lmod)
```

The $F$-statistic is $F=1.1398$ with corresponding $p$-value $=0.2963$.

#### Conclusion

The $p$-value is larger than just about any reasonable significance level, thus the test is not statistically significant and we fail to reject the claim that $\beta_{\rm Area} =0$. However, be careful, we do not accept the claim that $\beta_{\rm Area} =0$ either. We simply say the test is inconclusive.


Comparing the output above with full model test.

```{r}
summary(lmod)
```


## Using a t-statistic

An alternative approach is to use a $t$-statistic for testing the hypotheses:

$$ t_i = \frac{\widehat{\beta}_i-0}{\mbox{se}(\widehat{\beta}_i)}= \frac{\widehat{\beta}_i}{\mbox{se}(\widehat{\beta}_i)}$$
using a $t$-distribution with $\mbox{df}_{\Omega}=n-p$ degrees of freedom.

```{r}
# calculation of t-stat
beta.area <- summary(lmod)$coeff[2,1]
se.area <- summary(lmod)$coeff[2,2]
(t <- beta.area/se.area)

# verifying p-value
2*pt(t, df.residual(lmod))

# matching F-stat
t^2
```

## Question 13 {#sec-q13}

What is the difference between the test being run in the code below and the previous test? Does the test below lead to the same conclusion?

### Solution to Question 13

The test below is comparing the model with one regressor (`Area`) compared to the null (constant mean) model. We are not considering any other predictors.

- $H_0: \beta_1=\beta_{\rm Area}=0$

- $H_a: \beta_1 \neq 0$


```{r}
# compare results to test of beta_Area = 0 when no other predictors in model
summary(lm(Species ~ Area, gala))
```

According to the test above, it seems that the regression coefficient for `Area` is statistically significant and we do have evidence to support the claim that $\beta_{\rm Area} \ne 0$. This contradicts the previous test. Perhaps it is the other predictors in the full model that have an effect on `Area` that in turn is related to a change in the response. Thus `Area` may be a confounding variable and the other predictors may be responsible for this response in both `Area` and `Species`


# Testing a Pair of Regressors

## Question 14 {#sec-q14}

Test whether the `Area` and `Adjacent` regressor variables should be simultaneously dropped from the model that already includes `Elevation`, `Nearest`, and `Scruz` in the model.  Make sure to specify the regressors that are the model when stating H~0~ and H~a~.

- State the hypotheses. 
- Compute the $F$-statistic and $p$-value.
- State the conclusion in this context.


### Solution to Question 14

#### Hypotheses

- $H_0: \beta_{\rm Area}=\beta_{\rm Adjacent} 0 \mid \beta_0, \beta_2, \beta_3, \beta_4 \in \mathbb{R}$

- $H_a: \beta_{\rm Area} \ne 0 \mbox{ or }  \beta_{\rm Adjacent} \ne 0 \mid \beta_0, \beta_2, \beta_3, \beta_4 \in \mathbb{R}$


#### Compute the $F$-statistic and $p$-value

```{r}
# Determine full and reduced models
lmods <- lm(Species ~ Elevation + Nearest + Scruz, data = gala) # fit reduced model
anova(lmods, lmod) # compare models using general f-test
```

The $F$-statistic is $F=9.2874$ with corresponding $p$-value $=0.0.0013$.

#### Conclusion in this context.

The $p$-value $<0.001$ which is extremely small (less than just about any significance level that would be chosen). The test is statistically significant. There is very strong evidence that the regression model for `Species` that includes predictors `Area` and `Adjacent` is preferable to the model without both of these terms. We have evidence that at least one of $\beta_{\rm Area}$ and/or $\beta_{\rm Adjacent}$ are nonzero.


# Appendix A: Summary of Hypothesis Testing {#sec-appendA}

1.  State the  [**hypotheses**]{style="color:blue"} and identify (from the alternative claim in $H_a$) if it is a one or two-tailed test.

-   $H_0$ is the “boring” claim. Express using an equal sign $=$.
-   $H_a$ is the claim we want to show is likely true. Use inequality sign ($>$, $<$, or $\ne$).
-   State both $H_0$ and $H_a$ in terms of population parameters such as $\mu$ and $p$.

2.  Compute the  [**test statistic**]{style="color:blue"}.

-   If the observed sample contradicts the null claim, the result is significant.
-   A standardized test statistic measures how many SE's the observed stat is from the null claim.
-   A standardized test statistic with a large absolute value is supporting evidence to reject $H_0$.

3.  Using the null distribution, compute the  [**p-value**]{style="color:blue"}. The p-value is the probability of getting a sample with a test statistic as or more extreme than the observed sample assuming $H_0$ is true.

-   The p-value is the area in one or both tails beyond the test statistic.
-   The p-value is a probability, so we have $0 < \mbox{p-value} < 1$.
-   The smaller the p-value, the stronger the evidence to reject $H_0$.

4.  Based on the  [**significance level**]{style="color:blue"}, $\alpha$, make a decision to reject or not reject the null hypothesis

-   If p-value $\leq \alpha$, we reject $H_0$.
-   If p-value $> \alpha$, we do not reject $H_0$.

5. [**Summarize the results**]{style="color:blue"} in practical terms, **in the context of the example**.

-   If we reject $H_0$, this means there is enough evidence to support the claim in $H_a$.
-   If we do not reject $H_0$, this means there is not evidence to reject $H_0$ nor support $H_a$. The test is inconclusive.

# Appendix B: Deriving the Standard Error


