---
title: "Solutions 8: Estimation with Multiple Linear Regression"
author: 
  - name: "Adam Spiegler"
    affiliation: "University of Colorado Denver"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    toc-expand: 1
    embed-resources: true
editor: source
#execute:
#  eval: false
title-block-banner: true
warning: false
license: "CC BY-NC-SA"
---


## Additional Reading

---

- [Chapter 3 Linear Model Estimation](https://jfrench.github.io/LinearRegression/linear-model-estimation.html) from *A Progressive Introduction to Linear Models*
- Chapter 4 of *Linear Regression Models: Applications in R* ([link to library ebook](https://go.exlibris.link/69jsQwWW))
- Chapter 3 of *Applied Linear Regression* ([link to library ebook](https://go.exlibris.link/69jsQwWW))


# Loading Required Packages
 
---

State assumptions (again) and state in terms of matrices after matrix notation introduced. See Chap 3 Applied Linear Regression.

```{r}
library(faraway)  # to access Galapagos
library(tidyverse)
library(microbenchmark)  # for comparing computation speeds
```


# Multiple Linear Regression Models


A [**multiple linear regression**]{style="color:blue"} model is a linear regression model with more than one predictor $X_i$ and a single response variable $Y$. In general, we include an intercept term as a predictor, so if there are $p$ predictor variables, our model has $p+1$ regression coefficients. The most common multiple linear regression model is

$$Y = \beta_0 + \beta_1 X_1  + \beta_2 X_2 + \ldots + \beta_p X_p + \epsilon .$$

The model does not need to be linear in the predictor variables. For example, we can construct a polynomial with one predictor using multiple linear regression:

$$Y = \beta_0 + \beta_1 X  + \beta_2 X^2 + \ldots + \beta_p X^p + \epsilon .$$

We can transform some predictor variables and the response in our model. For example,

$$\log{(Y)} = \beta_0 + \beta_1 X_1  + \beta_2 \log{(X_2)} + \beta_3 X_3^2 + \epsilon .$$
Combining several predictors into a single term in the model. Products of regressors are called [**interactions**]{style="color:blue"}. They are often included to allow for joint effects of predictors.

$$Y = \beta_0 + \beta_1 X_1  + \beta_2 X_2 + \beta_3 X_1X_2 + \epsilon .$$

## Random Vectors


The notation $Y$ and $\epsilon$ represent [**random vectors**]{style="color:blue"}. Suppose we have a sample on $n$ observations. Then for each set of observed values, we have two $n\times1$ vectors of random variables:

$$\mathbf{y} = \left[  \begin{array}{c}  Y_1 \\ Y_2 \\ \vdots \\ Y_n \end{array}  \right] 
\qquad \text{and} \qquad 
\boldsymbol{\epsilon} = \left[  \begin{array}{c}  \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_n \end{array}  \right]$$

We use lower-case $\mathbf{y}$ since bold lower-case letters are typically used to denote vectors in mathematics while bold upper-case letters typically represent matrices. See [Appendix](#sec-appendix) for information about finding expected value, variance, and covariance of random vectors.

::: {.callout-note}
We typically will not denote random vectors using bold font and instead will use regular font $Y$ and $\epsilon$ to denote random vectors and $Y_i$ and $\epsilon_i$ to denote the $i^{\text{th}}$ entries in each vector. We also may often express the vectors as the transpose of $1 \times n$ vectors since that saves vertical space on the page.

$$y = \left[   Y_1 , Y_2 ,\ldots , Y_n \right]^T
\ \ \text{ and }  \ \ 
\epsilon = \left[  \epsilon_1 \\ \epsilon_2 , \ldots, \epsilon_n \right]^T$$
:::





# The Ingredients of Multiple Linear Regression

We now consider the context where we want to estimate the parameters of a linear model with 1 or more regressors, i.e., when 

$$ Y=\beta_0 + \beta_1 X_1 + \cdots + \beta_{p-1} X_{p-1} + \epsilon.$$

We define the following notation before expressing the model as an equivalent matrix equation.

- $\mathbf{y} = \lbrack Y_1, Y_2, \ldots, Y_n \rbrack^T$ denotes the column vector containing the $n$ response variables.
  - It is am $n \times 1$ random vector whose entries are each a random variable $Y_i$.
  - For each observation, we can observe a value for the response variable $y_i$, but it varies.
  - For each observation, we can compute the fitted value $\hat{y}_i$ to predict the value of $Y_i$.
  - We denote $\mathbf{\hat{y}}$ the $n \times 1$ vector of fitted values which we can compute and observe.

- $\mathbf{X}$ denotes the matrix containing a column of 1's (for the intercept) followed by columns of values corresponding to each regressor in our model.
  - The entries in column 2 ($x_{11}, x_{21}, \ldots , x_{n1}$) correspond to the values of regressor variable $X_1$. $x_{i1}$ is the value of $X_1$  for the $i^{\text{th}}$ observation.
  - $\mathbf{X}$ consists fixed and observable values.

$$ \mathbf{X} = \begin{bmatrix}
1 & x_{11} & x_{12} & \cdots & x_{1,p-1} \\
1 & x_{21} & x_{22} & \cdots & x_{2,p-1} \\
\vdots & \vdots & \vdots & \vdots & \vdots \\
1 & x_{n1} & x_{n2} & \cdots & x_{n,p-1}
\end{bmatrix}.$$


- $\boldsymbol{\beta} = [\beta_0, \beta_1, \ldots, \beta_{p-1}]^T$ denotes the $p \times 1$ column vector containing the $p$ regression coefficients.
  - The $\boldsymbol{\beta}$ is parameter whose value is fixed (not random) but unknown (not observable).
  - We can use different methods to find estimators $\boldsymbol{\hat{\beta}} = [\hat{\beta}_0, \hat{\beta}_1, \ldots, \hat{\beta}_{p-1}]^T$.
  - We will use OLS estimates computed by minimizing the RSS.

- $\boldsymbol{\epsilon} = [\epsilon_1, \epsilon_2, \ldots, \epsilon_n]^T$ denotes the $n \times 1$ random vector containing random variables corresponding to the $n$ errors.
  - Recall each $\epsilon_i$ is a random variable.
  - We can estimate the error using an $n \times 1$ vector of observed residuals, $\boldsymbol{\hat{\epsilon}}$.


Then the system of equations defining the linear model can be written as a matrix equation :

$$\large{ \mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}. }$$

::: {.callout-note}
Each $Y_i$ is a random variable since with different samples, we may observe different values of $Y_i$ given the same value(s) for the predictor(s). In general, if the object's value varies due to randomness in sampling, it is a random variable. 
:::


## Question 1


Summarize this information in the table below for the objects previously discussed using the various notations introduced. 

### Solution to Question 1


| Notation |  Description | Observable | Random |
|----------|--------------|------------|--------|
| $Y$ | response variable | Yes | Yes | 
| $Y_i$ | response value for the $i^{\text{th}}$ observation (we observe one value out of the many possible values) | Yes | Yes | 
| $\mathbf{y}$ | the $n\times 1$ column vector of response values | Yes | Yes | 
| $X_j$ | the $j$th predictor variable | Yes | No | 
| $x_{i,j}$ | the value of the $j$th predictor variable for the $i$th observation | Yes | No | 
| $\mathbf{X}$ | the $n\times p$ matrix of predictor values | Yes | No | 
| $\beta_j$ | the regression coefficient associated with the $j$th predictor variable  | No | No | 
| $\boldsymbol{\beta}$ | the $p\times 1$ column vector of regression coefficients | No | No | 
| $\boldsymbol{\epsilon}$ | the $n\times 1$ column vector of errors | No | Yes | 
| $\epsilon_i$ | the error associated with observation $i$ | No | Yes | 
| $\hat{\beta}_j$ | the estimator for $\beta_j$ | Yes | Yes | 
| $\boldsymbol{\hat{\beta}}$ | the $p\times 1$ column vector of estimates for the regression coefficients | Yes | Yes | 
| $\hat{\epsilon}_i$ | the residual associated with the $i^{\text{th}}$ observation| Yes | Yes | 


# OLS Estimators


The least squares estimator of $\boldsymbol{\beta}$ , denoted $\boldsymbol{\hat{\beta}}$, minimizes 

$$ \text{RSS}(\boldsymbol{\hat{\beta}})= \sum \hat{\epsilon}_i^2 = \boldsymbol{\hat{\epsilon}}^T \boldsymbol{\hat{\epsilon}} = (\mathbf{y}  -\mathbf{X}\boldsymbol{\hat{\beta}})^T( \mathbf{y}  -\mathbf{X}\boldsymbol{\hat{\beta}}) .$$

[How do we find the value of $\boldsymbol{\hat{\beta}}$ that minimizes RSS?]{style="color:blue"}

- With two predictors, we want to find the plane that minimizes the sum of squared distances (projections) of each point to the plane (regression line).
- Same idea in higher dimensions with more predictors.

```{r}
#| eval: true
#| echo: false
#| warning: false
library(plotly)
library(reshape2) # for acast()

country <- read_csv("https://raw.githubusercontent.com/aspiegler/Applied-Regression/main/Data/country.csv")

country <- country |> 
  mutate(log_gdp = log(gdp_per_cap), .before = gdp_per_cap) |>
  select(-gdp_per_cap)  # remove gdp_per_cap

model <- lm(life_exp ~ log_gdp + fertility, data = country)  # create model
x_grid <- seq(min(country$log_gdp), max(country$log_gdp), length.out = 50)  # grid for x
y_grid <- seq(min(country$fertility), max(country$fertility), length.out = 50)  # grid for y
regression_grid <- expand.grid(log_gdp = x_grid, fertility = y_grid)  # grid for z
regression_grid$life_exp_pred <- predict(model, newdata = regression_grid)  # fitted points
z_matrix <- acast(regression_grid, fertility ~ log_gdp, value.var = "life_exp_pred")  # merge vectors

plot_ly(country, x = ~log_gdp, y = ~fertility, z = ~life_exp,
        type = "scatter3d", mode = "markers", name = "Data Points") |> 
  add_surface(x = ~x_grid, y = ~y_grid, z = ~z_matrix,
              name = "Regression Plane",
              opacity = 0.5) |>
  layout(
    scene = list(
      xaxis = list(title = "log(GDP)"),
      yaxis = list(title = "Fertility"),
      zaxis = list(title = "Life Expectancy")
    )
  )
```



## Matrix Derivatives 

Now lets look at the derivative of a scalar function with respect to a vector.


Let $f(\mathbf{x})$ be a function of a $p\times 1$ column vector $\mathbf{x}=[x_1, x_2, \ldots,  x_p]^T$. The derivative of $f(\mathbf{x})$ with respect to $\mathbf{x}$ is denoted $\frac{\partial f(\mathbf{x})}{\partial \mathbf{x}}$ and 

$$\frac{\partial f(\mathbf{x})}{\partial \mathbf{x}} = \begin{bmatrix}
\frac{\partial f(\mathbf{x})}{\partial x_1}\\
\frac{\partial f(\mathbf{x})}{\partial x_2}\\
\vdots \\
\frac{\partial f(\mathbf{x})}{\partial x_p}
\end{bmatrix}.$$


Let

$$\mathbf{x} = \begin{bmatrix} x_1 \\ x_2\\  x_3 \end{bmatrix}, \quad
\mathbf{a} = \begin{bmatrix} a_1 \\ a_2\\  a_3 \end{bmatrix}, \quad \text{and }
\mathbf{A} = \begin{bmatrix} a_{11} & a_{12} \\
a_{21} & a_{22} \\
a_{31} & a_{32} \end{bmatrix},$$

and consider the examples below.

::: {.callout-note title="Example 1" appearance="simple" collapse="true"}
$f(\mathbf{x}) = \mathbf{x}^T \mathbf{a} = a_1x_1 + a_2x_2 + a_3x_3$ has the following derivative:

$$\frac{d}{d \mathbf{x}} \left( \mathbf{x}^T \mathbf{a} \right) = \begin{bmatrix}
 \frac{d}{dx_1} (a_1x_1 + a_2x_2 + a_3b_3) \\
 \frac{d}{dx_2} (a_1x_1 + a_2x_2 + a_3b_3) \\
 \frac{d}{dx_3} (a_1x_1 + a_2x_2 + a_3b_3) \end{bmatrix} = \begin{bmatrix}
 a_1 \\ a_2 \\ a_3 
 \end{bmatrix} = \mathbf{a}
 $$

More generally (for any dimension) $\frac{d}{d \mathbf{x}} \left( \mathbf{x}^T \mathbf{a} \right) = \mathbf{a}$.
:::



::: {.callout-note title="Example 2" appearance="simple" collapse="true"}
$f(\mathbf{x}) = \mathbf{x}^T \mathbf{x} = x_1^2 + x_2^2 + x_3^2$ has the following derivative:

$$\frac{d}{d \mathbf{x}} \left( \mathbf{x}^T \mathbf{x} \right) = \begin{bmatrix}
 \frac{d}{dx_1} (x_1^2 + x_2^2 + x_3^2) \\
 \frac{d}{dx_2} (x_1^2 + x_2^2 + x_3^2) \\
 \frac{d}{dx_3} (x_1^2 + x_2^2 + x_3^2) \end{bmatrix} = \begin{bmatrix}
 2x_1 \\ 2x_2 \\ 2x_3 
 \end{bmatrix} = 2 \mathbf{x}
 $$

More generally (for any dimension) $\frac{d}{d \mathbf{x}} \left( \mathbf{x}^T \mathbf{x} \right) = 2\mathbf{x}$.
:::


::: {.callout-note title="Example 3" appearance="simple" collapse="true"}
Let $\mathbf{a}_1$ and $\mathbf{a}_2$ denote the column vectors of $3 \times 2$ matrix $\mathbf{A}$ defined above. Consider the function

$$f(\mathbf{x}) = \mathbf{x}^T \mathbf{A} = \begin{bmatrix} x_1 & x_2 & x_3 \end{bmatrix} \begin{bmatrix} a_{11} & a_{12} \\
a_{21} & a_{22} \\
a_{31} & a_{32} \end{bmatrix} = \begin{bmatrix} \mathbf{x}^T \mathbf{a}_1 & \mathbf{x}^T \mathbf{a}_2 \end{bmatrix} .$$



$$\begin{align}
\frac{d}{d \mathbf{x}} \left( \mathbf{x}^T \mathbf{A} \right) &= 
\frac{d}{d \mathbf{x}} \left( \begin{bmatrix} \mathbf{x}^T \mathbf{a}_1 & \mathbf{x}^T \mathbf{a}_2 \end{bmatrix} \right) \\
&=\begin{bmatrix} \frac{d}{d \mathbf{x}} \left( \mathbf{x}^T \mathbf{a}_1 \right) & \frac{d}{d \mathbf{x}} \left( \mathbf{x}^T \mathbf{a}_2 \right) \end{bmatrix}  \\
&= \begin{bmatrix} \mathbf{a}_1 & \mathbf{a}_2 \end{bmatrix} \\
&= \mathbf{A}
 \end{align}
 $$

We can similarly show, for appropriately sized matrix $\mathbf{A}$,

$$\frac{d}{d \mathbf{x}} \left( \mathbf{A} \mathbf{x} \right) = \mathbf{A}$$
:::




## Question 2

Find a formula for OLS estimator for $\boldsymbol{\epsilon}$ using the following method:

- Expand the product $(\mathbf{y}  -\mathbf{X}\boldsymbol{\hat{\beta}})^T( \mathbf{y}  -\mathbf{X}\boldsymbol{\hat{\beta}})$. 
- Then take the derivative with respect to $\boldsymbol{\hat{\beta}}$ and set equal to $\mathbf{0}$.
- Solve for $\boldsymbol{\hat{\beta}}$. 


### Solution to Question 2


#### Expand the product 


$$\begin{align}
(\mathbf{y}  -\mathbf{X}\boldsymbol{\hat{\beta}})^T( \mathbf{y}  -\mathbf{X}\boldsymbol{\hat{\beta}}) &= 
\big( \mathbf{y}^T  - (\mathbf{X}\boldsymbol{\hat{\beta}})^T \big) ( \mathbf{y}  -\mathbf{X}\boldsymbol{\hat{\beta}}) \\
 &= 
\big( \mathbf{y}^T  - \boldsymbol{\hat{\beta}}^T \mathbf{X}^T \big) ( \mathbf{y}  -\mathbf{X}\boldsymbol{\hat{\beta}}) \\
&= \mathbf{y}^T \mathbf{y} - \mathbf{y}^T \mathbf{X}\boldsymbol{\hat{\beta}} - \boldsymbol{\hat{\beta}}^T \mathbf{X}^T \mathbf{y}  + \boldsymbol{\hat{\beta}}^T \mathbf{X}^T \mathbf{X}\boldsymbol{\hat{\beta}} \\
&= \mathbf{y}^T \mathbf{y} - \mathbf{y}^T \mathbf{X}\boldsymbol{\hat{\beta}} - (\mathbf{X} \boldsymbol{\hat{\beta}})^T  \mathbf{y}  + (\mathbf{X}\boldsymbol{\hat{\beta}} )^T (\mathbf{X}\boldsymbol{\hat{\beta}} ) \\
&= \mathbf{y}^T \mathbf{y} - \mathbf{y}^T \mathbf{X}\boldsymbol{\hat{\beta}} -   \mathbf{y}^T\mathbf{X}\boldsymbol{\hat{\beta}}  +  (\mathbf{X}\boldsymbol{\hat{\beta}} )^T (\mathbf{X}\boldsymbol{\hat{\beta}} )\\
&= \mathbf{y}^T \mathbf{y} - 2\mathbf{y}^T \mathbf{X}\boldsymbol{\hat{\beta}}  +  (\mathbf{X}\boldsymbol{\hat{\beta}} )^T (\mathbf{X}\boldsymbol{\hat{\beta}} )
\end{align}$$. 

::: {.callout-caution}
In the work above (and below), note $(\mathbf{X} \boldsymbol{\hat{\beta}})^T  \mathbf{y}$ results in a scalar. Thus, taking the transpose again would result in the same scalar.

$$(\mathbf{X} \boldsymbol{\hat{\beta}})^T  \mathbf{y}  = \left( (\mathbf{X} \boldsymbol{\hat{\beta}})^T  \mathbf{y} \right)^T = \mathbf{y}^T (\mathbf{X} \boldsymbol{\hat{\beta}})$$

It is generally NOT true that  $\mathbf{A}^T \mathbf{B} = \mathbf{B}^T \mathbf{A}$ when working with appropriately sized matrices.
:::


#### Take the derivative with respect to $\boldsymbol{\hat{\beta}}$


$$\begin{align}
\frac{d}{d \boldsymbol{\hat{\beta}}} \left( \mathbf{y}^T \mathbf{y} - 2\mathbf{y}^T \mathbf{X}\boldsymbol{\hat{\beta}}   + (\mathbf{X}\boldsymbol{\hat{\beta}} )^T (\mathbf{X}\boldsymbol{\hat{\beta}} ) \right) &= 
\frac{d}{d \boldsymbol{\hat{\beta}}} \left( \mathbf{y}^T \mathbf{y} \right) - 2 \frac{d}{d \boldsymbol{\hat{\beta}}} \left( \mathbf{y}^T \mathbf{X}\boldsymbol{\hat{\beta}} \right) + \frac{d}{d \boldsymbol{\hat{\beta}}} \left( (\mathbf{X}\boldsymbol{\hat{\beta}} )^T (\mathbf{X}\boldsymbol{\hat{\beta}} ) \right) \\
&= 0 -2 \mathbf{y}^T \mathbf{X} + \frac{d}{d \boldsymbol{\hat{\beta}}} \left[  (\mathbf{X}\boldsymbol{\hat{\beta}} )^T \right] \left( \mathbf{X}\boldsymbol{\hat{\beta}} \right) +  \left( \mathbf{X}\boldsymbol{\hat{\beta}}  \right)^T \frac{d}{d \boldsymbol{\hat{\beta}}} \left[ (\mathbf{X}\boldsymbol{\hat{\beta}} ) \right] \\
&= -2 \mathbf{y}^T \mathbf{X} + \bigg( \frac{d}{d \boldsymbol{\hat{\beta}}} \left[  \mathbf{X}\boldsymbol{\hat{\beta}}  \right] \bigg)^T \left( \mathbf{X}\boldsymbol{\hat{\beta}} \right) +  \left( \mathbf{X}\boldsymbol{\hat{\beta}}  \right)^T \frac{d}{d \boldsymbol{\hat{\beta}}} \left[ (\mathbf{X}\boldsymbol{\hat{\beta}} ) \right] \\
&= -2 \mathbf{y}^T \mathbf{X} + (\mathbf{X})^T \left( \mathbf{X}\boldsymbol{\hat{\beta}} \right) +  \left( \mathbf{X}\boldsymbol{\hat{\beta}}  \right)^T  (\mathbf{X}) \\
&= -2 \mathbf{y}^T \mathbf{X} + 2 \mathbf{X}^T \mathbf{X}\boldsymbol{\hat{\beta}}
\end{align}$$. 


::: {.callout-caution}
Again we note a similar special case:

- $\mathbf{X}^T\mathbf{X} \boldsymbol{\hat{\beta}}$ results in a column vector with the same dimensions as $\boldsymbol{\hat{\beta}}$.
- $\boldsymbol{\hat{\beta}}^T \mathbf{X}^T \mathbf{X}$ results in the same vector as the product above, expressed in the form of a row vector.
- In order to add the vectors, we treat all as column vectors, and thus 

$$\mathbf{X}^T \mathbf{X}\boldsymbol{\hat{\beta}}= \boldsymbol{\hat{\beta}}^T \mathbf{X}^T  \mathbf{X} .$$
Want to be clear (again) that it is generally NOT true that  $\mathbf{A}^T \mathbf{B} = \mathbf{B}^T \mathbf{A}$ when working with appropriately sized matrices.
:::


#### When is derivative equal to zero?

To find critical points where the RSS is minimized, we set the derivative equal to $\mathbf{0}$ (the $n \times 1$ zero vector) and solve for $\boldsymbol{\hat{\beta}}$.
 
We have

$$-2 \mathbf{y}^T \mathbf{X} + 2 \mathbf{X}^T \mathbf{X}\boldsymbol{\hat{\beta}} = \mathbf{0}$$
which gives

$$\mathbf{X}^T \mathbf{X}\boldsymbol{\hat{\beta}} =  \mathbf{y}^T \mathbf{X}$$

or 

$$\mathbf{X}^T \mathbf{X}\boldsymbol{\hat{\beta}} =   \mathbf{X}^T \mathbf{y}.$$

::: {.callout-note}
The resulting system above is called the normal equations.
:::


Multiplying both sides (on the left) by $(X^TX)^{-1}$ (if it exists), gives the OLS estimator

$$\large{ \boldsymbol{\hat{\beta}} = \left( \mathbf{X}^T \mathbf{X} \right)^{-1}  \mathbf{X}^T \mathbf{y} }$$
where $\mathbf{X}$ is the matrix of observed values of the predictor variables and  $\mathbf{y}$ is the vector of observed values of the response variable. Since we know all of these values, we can compute $\boldsymbol{\hat{\beta}}$.

<br>  
<br>  


# We Stopped Here on Wednesday


# Appendix A: A Glossary of Useful Functions


## Basics of Linear Algebra in R

Some useful commands when working with matrices in R:

- `a %*% b` computes matrix product  where $a$ and $b$ denote matrices.
- `t(a)` gives the transpose of matrix $a$.
- `solve(a)` finds the inverse of matrix $a$.
- `solve(a,b)` solves the equation $ax=b$.
- `det(a)` computes determinant of square matrix $a$.
- `crossprod(x,y)` Computes the matrix cross product $x \times y = x^Ty$.
- `matrix()` can be used to define a matrix from scratch.
- `diag(n)` creates $\mathbf{I}_n$, the $n \times n$ identify matrix.


## Refresher for `lm()`

Some useful commands working with `lm`:

- `coef(lmod)` Returns only the values of the regression coefficients, $\boldsymbol{\hat{\beta}}$.
- `x <- model.matrix( ~ x1 + x2 + x3 + x4, data = data.frame)` creates matrix $X$.
- `y <- data.frame$response_var` creates vector of observed values of $\mathbf{y}$.
- `names(lmod)` Lists the quantities we can extract from the `lmod`.
- `e <- residuals(lmod)` creates the vector of residuals $\hat{\boldsymbol\epsilon}$.
- `deviance(lmod)` computes the RSS of the linear model.
- `df.residual(lmod)` computes the degrees of freedom $n-p$.
- `sigma(lmod)` computes the estimate for the standard error $\hat{\sigma}$.
- `vcov(lmod)` gives the estimated covariance matrix of the regression coefficients.



# Appendix B: Random Vectors {#sec-appendixb}

Let $\mathbf{y}=(Y_1,Y_2,\dots,Y_n )^T$ is a random vector if it is an $n\times1$ vector of random variables. 


## Mean, Variance, and Covariance

The mean of a random vector is

$$E(\mathbf{y})=\begin{pmatrix}E(Y_1)\\E(Y_2)\\\vdots\\E(Y_n)\end{pmatrix}.$$

The variance (covariance) of a random vector is

$$\begin{align}
\mathrm{Var}(\mathbf{y}) &= E(\mathbf{y}\mathbf{y}^T )-E(\mathbf{y})E(\mathbf{y})^T\\
\\
&= \begin{pmatrix}\mathrm{Var}(Y_1) & \mathrm{Cov}(Y_1,Y_2) &\dots &\mathrm{Cov}(Y_1,Y_n)\\\mathrm{Cov}(Y_2,Y_1 )&\mathrm{Var}(Y_2)&\dots&\mathrm{Cov}(Y_2,Y_n)\\\vdots&\vdots&\vdots&\vdots\\
\mathrm{Cov}(Y_n,Y_1)&\mathrm{Cov}(Y_n,Y_2)&\dots&\mathrm{Var}(Y_n)\end{pmatrix}\end{align}.$$

Let $\mathbf{x} = (X_1, X_2, \ldots, X_n)^T$ and $\mathbf{y} = (Y_1, Y_2, \ldots, Y_n)^T$ be $n\times 1$ random vectors.

The covariance between two random vectors is 
$$\mathrm{Cov}(\mathbf{x}, \mathbf{y}) = E(\mathbf{x} \mathbf{y}^T) - E(\mathbf{x}) E(\mathbf{y})^T.$$

## Properties of Transformations of Random Vectors

Define:

- $\mathbf{a}$ to be  an $n\times 1$ vector of constants
- $A$ to be an $m\times n$ matrix of constants
- $\mathbf{x}=(X_1,X_2,\ldots,X_n)^T$ to be an $n\times 1$ random vector
- $\mathbf{y}=(Y_1,Y_2,\ldots,Y_n)^T$ to be an $n\times 1$ random vector
-	$\mathbf{z}=(Z_1,Z_2,\ldots,Z_n)^T$ to be an $n\times 1$ random vector
- $0_{n\times n}$ to be an $n\times n$ matrix of zeros.

Then:

-	$E(A\mathbf{y})=AE(\mathbf{y}), E(\mathbf{y}A^T )=E(\mathbf{y}) A^T.$
-	$E(\mathbf{x}+\mathbf{y})=E(\mathbf{x})+E(\mathbf{y})$
-	$\mathrm{Var}(A\mathbf{y})=A \ \mathrm{Var}(\mathbf{y}) A^T$
-	$\mathrm{Cov}(\mathbf{x}+\mathbf{y},\mathbf{z})=\mathrm{Cov}(\mathbf{x},\mathbf{z})+\mathrm{Cov}(\mathbf{y},\mathbf{z})$
-	$\mathrm{Cov}(\mathbf{x},\mathbf{y}+\mathbf{z})=\mathrm{Cov}(\mathbf{x},\mathbf{y})+\mathrm{Cov}(\mathbf{x},\mathbf{z})$
-	$\mathrm{Cov}(A\mathbf{x},\mathbf{y})=A\ \mathrm{Cov}(\mathbf{x},\mathbf{y})$
- $\mathrm{Cov}(\mathbf{x},A\mathbf{y})=\mathrm{Cov}(\mathbf{x},\mathbf{y}) A^T$
- $\mathrm{Var}(a)= 0_{n\times n}$
- $\mathrm{Cov}(\mathbf{a},\mathbf{y})=0_{n\times n}$
- $\mathrm{Var}(\mathbf{a}+\mathbf{y})= \mathrm{Var}(\mathbf{y})$

