---
title: "Solutions 12: Inference"
author: 
  - name: "Adam Spiegler"
    affiliation: "University of Colorado Denver"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    toc-expand: 1
    embed-resources: true
editor: source
#execute:
#  eval: false
title-block-banner: true
warning: false
license: "CC BY-NC-SA"
---


## Additional Reading


- [Chapter 6 Linear Model Inference and Prediction](https://jfrench.github.io/LinearRegression/inference.html#overview-of-inference-and-prediction) from *A Progressive Introduction to Linear Models*
- Chapter 6 of *Linear Regression Models: Applications in R* ([link to library ebook](https://go.exlibris.link/Q7wgCRrz))
- Chapter 6 of *Applied Linear Regression* ([link to library ebook](https://go.exlibris.link/69jsQwWW))


# Loading Required Packages and Data


```{r}
#| message: false
library(tidyverse)
library(ellipse)
library(faraway)  # access gala data frame
```


::: {.callout-note}
We have not used the `ellipse` package yet this semester. If you have not installed this package, be sure to run the command `install.packages("ellipse")` from the console below and re-run the code cell above.
:::

```{r}
data(gala)
head(gala)
```

```{r}
summary(gala)
```

The data set `gala` contains 30 observations (one for each island in the Galapagos). There are 6 variables in the data set. The relationship between the number of plant species and several geographic variables is of interest.

- `Species`: The number of plant species found on the island.
- `Area`: The area of the island ($\mbox{km}^2$).
- `Elevation`: The highest elevation of the island (m).
- `Nearest`: The distance from the nearest island (km).
- `Scruz`: The distance from Santa Cruz Island (km).
- `Adjacent`: The area of the adjacent island ($\mbox{km}^2$).

# Hypothesis Testing for a Simple Linear Regression


## Question 1  {#sec-q1}

Write out a formula for the fitted model based on code and output below.

```{r}
lmod_area <- lm(Species ~ Area, data = gala)
summary(lmod_area)
```


### Solution to Question 1



$$\widehat{\mbox{Species}} = 63.78286 +  0.08196 \mbox{Area}$$


## Setting Up The Hypotheses


## Question 2 {#sec-q2}

What do you think would be suitable hypothesis to test whether the predictor `Area` should be included?

### Solution to Question 2

- H~0~: There is no linear association between the `Species` and `Area`. $\beta_1 =0$.
- H~a~: There is a linear association between the `Species` and `Area`. $\beta_1 \ne 0$.


## Calculating the Test Statistic

If the null hypotheses is true, how likely would we be to obtain the observed sample data?

$$t = \frac{\text{observed} - \text{null}}{\text{SE}} = \frac{\widehat{\beta}_1 - 0}{\text{SE}(\widehat{\beta}_1)}.$$ 

Recall the OLS estimate for $\hat{\beta}_1$ is

$$\hat{\beta}_1 = \frac{\sum (x_i - \bar{x})(y - \bar{y})}{\sum (x_i - \bar{x})^2}.$$


We can find $\text{SE}(\hat{\beta}_1) = \sqrt{\text{Var}(X)}$ which is given by the formula below:

$$
\text{SE}(\widehat{\beta}_1) = \sqrt{ \frac{\sum (\hat{y}_i - y_i)^2 / (n-2)}{\sum (x_i - \bar{x})^2 }}.
$$


- $\frac{\sum (\hat{y}_i - y_i)^2$ is the RSS of the model with `Area` as the only predictor.
- $n-2$ denotes the degrees of freedom with $2$ being the number of regression coefficients.
- $\sum (x_i - \bar{x})^2$ is the total sum of the squared deviance in the predictor.

## Question 3 {#sec-q3}

Approximate the standard error for $\widehat{\beta}_1$. Then use this result to compute the test statistic and p-value corresponding to this test.


### Solution to Question 3

```{r}
# standard error
n <- length(gala$Species)  # number of observations
x_bar <- mean(gala$Area)  # mean of the observed predictors
top <- deviance(lmod_area) / (n-2)  # RSS / (n-2)
bottom <- sum((gala$Area - x_bar)^2)  # sum (x_i - x-bar)^2
sqrt(top / bottom)
```


Thus, we have test statistic

$$t = \frac{\widehat{\beta}_1 - 0}{\text{SE}(\widehat{\beta}_1)} = \frac{0.08196 - 0}{0.01971} \approx  4.1583$$


```{r}
# compute test statistic
0.08196 / 0.01971
```


```{r}
# compute P-value
2 * pt(-4.158295, n - 2)
```


```{r}
summary(lmod_area)
```

Since the p-value $0.00027$ is less than the significance level chosen (5%), the test is statistically significant. We can reject $H_0$ and accept $H_a$. In this context we can say the linear association between `Area` and `Species` is significant, and the slope coefficient corresponding to `Area` is not equal to zero.

If we square the $t$ test statistic, we get 

```{r}
(0.08196 / 0.01971)^2
```

which is the value of the $F$-statistic stated in the summary output of the linear model.

```{r}
summary(lmod_area)
```

# Hypothesis Tests with Multiple Linear Regression

## Question 4  {#sec-q4}

Write out a formula for the fitted model based on code and output below.

```{r}
lmod <- lm(Species ~ Area + Elevation + Nearest +
            Scruz + Adjacent, data = gala)
summary(lmod)
```

```{r}
deviance(lmod)
```

```{r}
sum(residuals(lmod)^2)
```


```{r}
nullmod <- lm(Species ~ 1, data = gala)
summary(nullmod)
```

```{r}
deviance(nullmod)
```

### Solution to Question 4



$$\widehat{\mbox{Species}} = 7.068 -  0.0239 \mbox{Area} + 0.3195 \mbox{Elevation} + 0.0091 \mbox{Nearest} -0.2405 \mbox{Scruz} - 0.0748 \mbox{Adjacent}$$


## Question 5 {#sec-q5}

Which regressor has the strongest association with the number species on the island? Explain how you determined your answer.

### Solution to Question 5

- Hard to tell at this point. We can see the `Elevation` had the largest slope (in absolute terms) and the smallest p-value, so that seems important!

- The regressor `Scruz` has a large slope (in absolute terms), however the variability is quite large as indicated by the standard error, and thus we see it has a relatively large p-value, indicating that is plausible the coefficient is actually 0 or it could be positive.



# Setting Up The Hypotheses

How can we decide whether all or some of the regressor variables should be included in our model?


- Let $\Omega$ denote the proposed **complete or full model**.
- Let $\omega$ denote a **simpler model** that uses a subset of all regressors.

## Question 6 {#sec-q6}

What do you think would be suitable hypothesis to test whether $\Omega$ is a better model than $\omega$?

### Solution to Question 6

- H~0~: Simple Model $\omega$ is adequate.
- H~a~: Complete Model $\Omega$ is preferred.


# Measuring Significance

- If the models have a similar fit, then we prefer $\omega$ since it is simpler.
- If the model $\Omega$  is a much better model than $\omega$, then we prefer $\Omega$.
- If $\mbox{RSS}_{\omega} - \mbox{RSS}_{\Omega}$ is large, then $\Omega$ has a superior fit.
- How large of a difference is large enough?


## An Initial Test Statistic

$$\mbox{test stat} = \frac{\mbox{fit of observed model} - \mbox{fit of null model}}{\mbox{variability due to sampling}} = \frac{\mbox{RSS}_{\omega} - \mbox{RSS}_{\Omega}}{RSS_{\Omega}}.$$


## The $F$-Statistic


The [**$F$-statistic**]{style="color:blue"} compare the variance explained by the model to the variance not explained (the residuals). The $F$-statistic is used to test whether the **overall regression model is statistically significant**, not just individual predictors. Recall the following measurements related to a regression model:

- $\mbox{RSS} = \sum (Y_i - \widehat{Y}_i)^2$ is the [**Residual Sum of Squares (RSS)**]{style="color:blue"}.
- RSS measures the variability in our response that **is NOT explained by our model**.


Suppose that model $\Omega$ has $p$ estimated regression coefficients and model $\omega$ has $q$ estimated regression coefficients. In other words, model $\Omega$ considers adding $p-q$ predictors to simpler model $\omega$.

- $\mbox{RSS}_{\omega} - \mbox{RSS}_{\Omega}$ is the additional variance in the response that was not explained by $\omega$ (simpler model) but is explained by $\Omega$ (complete model).
- Dividing by $df_{\omega}-df_{\Omega} = (n - q) - (n - p) = p-q$ averages the change in the RSS per additional regressor added.
- The denominator $RSS_{\Omega}/df_{\Omega} = RSS_{\Omega}/(n-p)$ is the mean (per regressor) of the remaining unexplained variance.

$$F =  \frac{(\mbox{RSS}_{\omega} - \mbox{RSS}_{\Omega})/(df_{\omega}-df_{\Omega})}{RSS_{\Omega}/df_{\Omega}} =  \frac{(\mbox{RSS}_{\omega} - \mbox{RSS}_{\Omega})/(df_{\omega}-df_{\Omega})}{\hat{\sigma}_{\Omega}^2} \sim F_{p-q,n-p} .$$


The $F$-statistic is the ratio that compares the variance in the response explained by the complete model $\Omega$ to the variance not explained by the complete model. A **large $F$-statistic** indicates the **model's predictors are jointly useful** in explaining the response variable. 


::: {.callout-note}
We divide each $RSS$ by its degrees of freedom to calculate an **unbiased** estimate of the mean variance similar to how we divide by $n-1$ when finding an unbiased estimate for the standard deviation. This prevents underestimation.
:::


# Compute the p-value

If we assume $H_0$ is true, the $F$-statistic follows an $F$-distribution with degrees of freedom $p-q$ and $n-p$. 

## Visualizing the F-dist

```{r}
#| echo: false
#| message: false
#| warning: false
#library(tidyverse)

data.frame(f = 0:1000 / 100) %>% 
           mutate(df_25_26 = df(x = f, df1 = 25, df2 = 26),
                  df_05_26 = df(x = f, df1 = 5, df2 = 26),
                  df_05_10 = df(x = f, df1 = 5, df2 = 10)) %>%
#                  df_05_10 = df(x = f, df1 = 5, df2 = 10)) %>%
  gather(key = "df", value = "density", -f) %>%

  ggplot() +
  geom_line(aes(x = f, y = density, color = df, linetype=df)) +
  labs(title = "F at Various Degrees of Freedom",
       x = "F",
       y = "Density") + 
  xlim(0, 5) + 
  theme_bw()
```

- If $\mbox{df}_1 = p-1$ is constant and we increase $\mbox{df}_2 = \mbox{df}_{\Omega} = n-p$, the center hump is centered around the same $F$ value, but there is more area in the hump and less in the right tail.
- If $\mbox{df}_2 = n-p$ is constant and we increase $\mbox{df}_1 = \mbox{df}_{\Omega} = p-1$, the center hump shifts to the right and there is more area concentrated near the peak and less in the tail to the right.


Let `f_star` denote the $F$-statistic associated to our regression models, then we have

```
df1 <- p-q  # number of added predictors
df2 <- n-p  # degrees of freedom of complete model
1 - pf(f_star, df1, df2)  # area in upper tail
```

::: {.callout-note}
The $F$-statistic and $p$-value are computed when constructing a linear model with the `lm()` function in R. Both are provided in the `summary()` of an `lm` object. Thus, we typically just run the `lm()` command in R to obtain these values.
:::

# Make A Conclusion

Let $\alpha$ denote the [**significance level**]{style="color:blue"} of the test which is chosen by the researcher. By default, we use $\alpha = 0.05$. 

- If $p$-value $\leq \alpha$, the test is statistically significant.
  - We can reject $H_0$ and have significant evidence that $H_a$ is true.
  - The full model $\Omega$ is better than the simpler model $\omega$.

- If $p$-value $> \alpha$, the test is NOT statistically significant.
  - We fail to reject (or accept $H_0$) and cannot say anything about the claim in $H_a$.
  - The test is inconclusive.
  - We cannot be sure whether or not $\omega$ or $\Omega$ is a better model.
  - In which case, we typically use the simpler model $\omega$.



# Example Are Any Predictors Useful?

To see if any of the predictors in a possible model are useful, we compare the full model $\Omega$ given by $\mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\epsilon}$ to the null model $\omega$ which is the constant mean model $\mathbf{y} = \mu_Y + \boldsymbol{\epsilon}$ which we estimate with $\hat{y} = \bar{y}$.


## Question 7: Set up the hypotheses {#sec-q7}

Set up the hypotheses to see whether or not the full model in [Question 4](#sec-q4) is better model than the null model.

### Solution to Question 7

- $H_0$: The null model $\omega$ is adequate.  $\beta_1 = \beta_2 = \ldots = \beta_{p-1} = 0$.
- $H_a$: At least one of the regression coefficients is not zero.


## Question 8:  Measure Significance {#sec-q8}

Using the model from [Question 4](#sec-q4), identify the $F$-statistic and $p$-value for this test. 

*Bonus: Try to calculate the $F$-statistic using the formula above to confirm.*

### Solution to Question 8:

```{r}
summary(lmod)
```

The $F$-statistic is $15.7$ which has a $p$-value of $0.0000006838$. 

#### Confirming the `lm()` Output

The simpler model $\omega$ is now the null model with $q=1$.


```{r}
nullmod <- lm(Species ~ 1, data = gala)
summary(nullmod)
```

Since $\omega$ is the null model, in this case we have $\text{RSS}_{\omega} = TSS$ and $\text{TSS} - \text{RSS}_{\Omega} = SS_{\rm reg}$.

$$F =  \frac{(\mbox{RSS}_{\omega} - \mbox{RSS}_{\Omega})/(df_{\omega}-df_{\Omega})}{RSS_{\Omega}/df_{\Omega}} =  \frac{(\mbox{TSS} - \mbox{RSS}_{\Omega})/(p-1)}{\mbox{RSS}_{\Omega} /(n-p)} = \frac{\mbox{SS}_{\rm reg}/(p-1)}{\text{RSS}/(n-p)}.$$


```{r}
rss_cm <- deviance(lmod)  # RSS of complete model
tss <- deviance(nullmod)  # RSS of null model is called the TSS
ss_reg <- tss - rss_cm
n <- nrow(gala)  # number of observations in data
q <- 1  # null model has 1 regressor (intercept)
p <- 6  # complete model has 6 possible regressors
top_null <- ss_reg / (p - q)
bot_null <- rss_cm / (n - p)
(f_null <- top_null / bot_null)
```

The $F$-statistic is $15.69941$. Finally, we compute the $p$-value using $p-q=5$ and $n-p = 24$ degrees of freedom.

```{r}
1 - pf(f_null, p-q, n-p)
```

The $p$-value is $0.000000684$


## Question 9:  Make a conclusion (if possible) {#sec-q9}

Based on the previous results, state a formal conclusion of this test (if there is one). Then summary the conclusion in the context of this example.

### Solution to Question 9

Since the $p$-value $= 0.0000006838$ is less than any significance level that would be chosen, we can reject $H_0$ and we have significant evidence that $H_a$ is true. This means the regression model with predictors `Area`, `Elevation`, `Nearest`, `Scruz`, and `Adjacent` is a better model than the null model which would predict the mean number of species for every observation, no matter the values of `Area`, `Elevation`, `Nearest`, `Scruz`, and `Adjacent`.


# Analysis of Variation (ANOVA) Table

The information above is often presented in an [**analysis of variation (ANOVA) table**]{style="color:blue"} which is convenient for summarizing and organizing the various ingredients in the $F$-statistic. Below is a typical ANOVA table when comparing a full model $\Omega$ with $p$ regressors to a simpler model $\omega$ with $q$ regressors.


| Source | Deg. of Freedom | Sum of Squares | Mean Square | F |
|--------|-----------------|----------------|-------------|---|
| Regression | $p-q$ | $\text{RSS}_{\omega} - \text{RSS}_{\Omega}$  | $(\text{RSS}_{\omega} - \text{RSS}_{\Omega}) / (p-q)$ | $F$-stat |
| Residuals | $n-p$ | $\text{RSS}_{\Omega}$ | $\text{RSS}_{\Omega} / (n-p)$ | |
| Total | $n-q$ | $\text{RSS}_{\Omega}$ | | |


- **Regression**: The variance that is explained by the additional variable(s) of interest in $\Omega$ that are not included in $\omega$.
- **Residuals**: The variance that is not explained by the complete model $\Omega$. This is the $\text{RSS}_{\Omega}$.
- **Total**: The variation that is unexplained by the simpler model $\omega$.

There is the `anova()` function in R can be used compute components of the $F$-statistic corresponding to a test for a regression model. The command `anova(simple_lmod, complete_lmod)` gives an ANOVA table with the following organization.


| Source | Res.Df  | RSS | Df | Sum of Sq | F | Pr( > F) |
|--------|---------|-----|----|-----------|---|----------|
| 1 (simple model, $\omega$) | $n-q$ | $\text{RSS}_{\omega}$ |  |  |  |  |
| 2 (complete model, $\Omega$) | $n-p$ | $\text{RSS}_{\Omega}$ | $p-q$  | $\text{RSS}_{\omega} - \text{RSS}_{\Omega}$ | $F$-stat | $p$-value |


## Question 10 {#sec-q10} 

In questions 7-9 we performed a hypothesis test to compare the full model (include all 5 predictors stored in `lmod`) to the null model (intercept only stored in `nullmod`). Play around with the code below to fill in the ANOVA table provided below.

```{r}
nullmod <- lm(Species ~ 1, data = gala)
```


```{r}
deviance(lmod)
deviance(nullmod)
deviance(nullmod) - deviance(lmod)
n <- nrow(gala)  # number of observations
p <- 6  # 6 regressors in full model
q <- 1  # 1 regressor in null model
```


### Solution to Question 10


```{r}
top_q10 <- (deviance(nullmod) - deviance(lmod)) / (p-q)
bot_q10 <- deviance(lmod) / (n - p)
f_q10 <- top_q10 / bot_q10
top_q10
bot_q10
f_q10
```



| Source | Deg. of Freedom | Sum of Squares | Mean Square | F | 
|--------|-----------------|----------------|-------------|---|
| Regression | $5$ | $291850$ | $58370$ | $15.69941$ |
| Residuals | $24$ | $89231$ | $3717.974$ | |
| Total | $29$ | $381081.4$ | | |



## Question 11 {#sec-q11}

Run the code cell below to generate an ANOVA table in R and compare with the table you created in the previous question. Then run the second code cell to compare with the summary output of the complete model, `lmod`.


```{r}
anova(nullmod, lmod)
```

```{r}
summary(lmod)
```


### Solution to Question 11


Everything looks consistent!


# Testing a Single Regressor


To test whether a single regressor (regressor j) can be dropped from the model, we choose between $H_0:\beta_j = 0$ and $H_a: \beta_j \ne 0$.

We have two options in this case:

- Use the previous approach, letting the reduced model be the one without that regressor.
- Use a $t$-statistic approach.


## Question 12 {#sec-q12}

Repeat the previous process to test whether the regressor `Area` is significant (assuming all others are in the model).

- State the hypotheses.
- Compute the $F$-statistic and $p$-value.
- State the conclusion in this context.


### Solution to Question 12


#### Hypotheses

- $H_0: \beta_1=\beta_{\rm Area}=0 \mid \beta_0, \beta_2, \beta_3, \beta_4, \beta_5 \in \mathbb{R}$

- $H_a: \beta_1 \neq 0\mid \beta_0, \beta_2, \beta_3, \beta_4, \beta_5 \in \mathbb{R}$

#### Compute the $F$-statistic and $p$-value.

```{r}
# fit reduced model (full without Area)
lmods <- lm(Species ~ Elevation + Nearest + Scruz + Adjacent, data = gala)
anova(lmods, lmod)
```

The $F$-statistic is $F=1.1398$ with corresponding $p$-value $=0.2963$.

#### Conclusion

The $p$-value is larger than just about any reasonable significance level, thus the test is not statistically significant and we fail to reject the claim that $\beta_{\rm Area} =0$. However, be careful, we do not accept the claim that $\beta_{\rm Area} =0$ either. We simply say the test is inconclusive.


Comparing the output above with full model test.

```{r}
summary(lmod)
```


## Using a t-statistic

An alternative approach is to use a $t$-statistic for testing the hypotheses:

$$ t_i = \frac{\widehat{\beta}_i-0}{\mbox{se}(\widehat{\beta}_i)}= \frac{\widehat{\beta}_i}{\mbox{se}(\widehat{\beta}_i)}$$
using a $t$-distribution with $\mbox{df}_{\Omega}=n-p$ degrees of freedom.

```{r}
# calculation of t-stat
beta.area <- summary(lmod)$coeff[2,1]
se.area <- summary(lmod)$coeff[2,2]
(t <- beta.area/se.area)

# verifying p-value
2*pt(t, df.residual(lmod))

# matching F-stat
t^2
```

## Question 13 {#sec-q13}

What is the difference between the test being run in the code below and the previous test? Does the test below lead to the same conclusion?


```{r}
# compare results to test of beta_Area = 0 when no other predictors in model
summary(lm(Species ~ Area, gala))
```


### Solution to Question 13


The test below is comparing the model with one regressor (`Area`) compared to the null (constant mean) model. We are not considering any other predictors.

- $H_0: \beta_1=\beta_{\rm Area}=0$

- $H_a: \beta_1 \neq 0$


According to the test above, it seems that the regression coefficient for `Area` is statistically significant and we do have evidence to support the claim that $\beta_{\rm Area} \ne 0$. This contradicts the previous test. Perhaps it is the other predictors in the full model that have an effect on `Area` that in turn is related to a change in the response. Thus `Area` may be a confounding variable and the other predictors may be responsible for this response in both `Area` and `Species`


# Testing a Pair of Regressors

## Question 14 {#sec-q14}

Test whether the `Area` and `Adjacent` regressor variables should be simultaneously dropped from the model that already includes `Elevation`, `Nearest`, and `Scruz` in the model.  Make sure to specify the regressors that are the model when stating H~0~ and H~a~.

- State the hypotheses. 
- Compute the $F$-statistic and $p$-value.
- State the conclusion in this context.


### Solution to Question 14


#### Hypotheses

- $H_0: \beta_{\rm Area}=\beta_{\rm Adjacent} 0 \mid \beta_0, \beta_2, \beta_3, \beta_4 \in \mathbb{R}$

- $H_a: \beta_{\rm Area} \ne 0 \mbox{ or }  \beta_{\rm Adjacent} \ne 0 \mid \beta_0, \beta_2, \beta_3, \beta_4 \in \mathbb{R}$


#### Compute the $F$-statistic and $p$-value

```{r}
# Determine full and reduced models
lmods <- lm(Species ~ Elevation + Nearest + Scruz, data = gala) # fit reduced model
anova(lmods, lmod) # compare models using general f-test
```

The $F$-statistic is $F=9.2874$ with corresponding $p$-value $=0.0.0013$.

#### Conclusion in this context.

The $p$-value $<0.001$ which is extremely small (less than just about any significance level that would be chosen). The test is statistically significant. There is very strong evidence that the regression model for `Species` that includes predictors `Area` and `Adjacent` is preferable to the model without both of these terms. We have evidence that at least one of $\beta_{\rm Area}$ and/or $\beta_{\rm Adjacent}$ are nonzero.


# Confidence Intervals

An alternative way of expressing the uncertainty in an estimate for a population parameter is through [**confidence intervals (CIs)**]{style="color:blue"}. A confidence interval interval is an interval estimate with an associated confidence level corresponding to the success rate.


## Example: Confidence Interval for Mean

Suppose we want to find a 95% confidence interval to estimate the average weight (in pounds) of all Komodo dragons. We have a sample of $n=8$ Komodo dragons whose sample mean weight is $\bar{x} = 155.5$ pounds with a sample standard deviation of $s = 22.55153$ pounds.

```{r}
dragon_wt <- c(145, 178, 142, 139, 160, 190, 168, 122)  # load sample of weights  
(wt_mean <- mean(dragon_wt))
(wt_sd <- sd(dragon_wt))
wt_n <- length(dragon_wt)
```

- A reasonable **point estimate** is $\mu \approx \bar{x} = 155.5$.
- Another person with a slightly different sample we have a slightly different estimate.
- Let's build in a margin of error into our estimate to account for the variability due to the randomness of sampling.

A corresponding $t$-distribution confidence interval is given by

$${\large \overline{X} - {\color{red}{t_{n-1}^{\alpha/2}}} \cdot \frac{s}{\sqrt{n}} < \mu < \overline{X} + {\color{red}{t_{n-1}^{\alpha/2}}} \cdot \frac{s}{\sqrt{n}}},$$

where the area under the $t$-distribution with $n-1$ degrees of freedom between ${\color{red}{\pm t_{\alpha/2}}}$ is equal to the confidence level.


```{r}
se_wt <- wt_sd / sqrt(n)  # estimate for standard error of x-bar
```


```{r}
df_wt <- 8 - 1  # n - 1 degrees of freedom
(t_star <- qt(0.025, df_wt))  # exclude lower 2.5%
```


```{r}
(moe_wt <- abs(t_star) * se_wt)
(lower_wt <- wt_mean - moe_wt)
(upper_wt <- wt_mean + moe_wt)
```

A 95% confidence interval for the mean weight of all Komodo dragons is therefore $155.5 \pm 9.74$ pounds or

$$145.76 \text{ pounds} < \mu < 165.24 \text{ pounds}.$$

### Interpreting a 95% Confidence Interval

There is a 95% chance the interval from $145.76$ to $165.24$ pounds contains the actual mean weight of all adult Komodo dragons.

- The interval is used to estimate a population parameter. In this case, the mean of ALL adult Komodo dragons.
- We should indicate the odds of success.
- We should attach the uncertainty to the interval we constructed (which is random).
- We should NOT attach the uncertainty to the parameter we are estimating (which is not random, it is fixed).


## Confidence Intervals for Regression Coefficients

To construct a confidence interval for a regression coefficient with confidence level $CL = 1 - \alpha$, we have:

$$\widehat{\beta}_j \ \pm \  {\color{red}{t_{n-p}^{\alpha/2}}} \ \ \mbox{SE}(\widehat{\beta}_j).$$

The resulting interval has a given odds of success (confidence level) for containing the actual value of the regression coefficient.


## Question 15 {#sec-q15}

Use the output from the code below to constructing a 95% confidence interval for the `Area` slope coefficient $\beta_{\rm Area}$ (assuming all 4 other predictors in the model). Interpret the practical meaning of the interval.

```{r}
lmod <- lm(Species ~ Area + Elevation + Nearest + Scruz + Adjacent, data = gala)
summary(lmod)
```


### Solution to Question 15


```{r}
n_gala <- nrow(gala)  # number of observations
p <- 6  # number of regressors (includes intercept)
tstar <- qt(0.025, n_gala - 6)  # Find value of t_{n-p}
point_est_area <- -0.023938 
se_beta_area <- 0.022422
(cutoffs_beta_area <- point_est_area + c(-1,1) * tstar * se_beta_area)
```


There is a 95% chance the interval from $-0.0702$ to $0.0223$  contains the actual value of $\beta_{\rm Area}$.

::: {.callout-note}
Notice the 95% confidence interval for $\beta_{\rm Area}$ contains the value $0$. This means $0$ is a plausible value for $\beta_{\rm Area}$, and the linear relation between `Species` and `Area` in this model is not statistically significant at a 5% significance level. This is consistent with the analysis of the p-value $0.296$.
:::

## Question 16 {#sec-q16}

Use the summary output from the code above to similarly construct a 95% confidence interval for the `Elevation` slope coefficient (assuming all 4 other predictors in the model). Interpret the practical meaning of the interval.


### Solution to Question 16

```{r}
point_est_elav <- 0.319465
se_beta_elav <- 0.053663
(cutoffs_beta_elav <- point_est_elav + c(-1,1) * tstar * se_beta_elav)
```


There is a 95% chance the interval from $0.2087$ to $0.4302$ contains the actual value of $\beta_{\rm Elevation}$.

::: {.callout-note}
Notice the 95% confidence interval for $\beta_{\rm Elevation}$ does NOT contain the value $0$. This means $0$ is NOT a plausible value for $\beta_{\rm Elevation}$, and the linear relation between `Species` and `Elevation` in this model is statistically significant at the 5% significance level. This is consistent with the analysis of the p-value `3.82e-06`.
:::

## Finding Confidence Intervals for All Regressors

```{r}
confint(lmod)
```



# Confidence Regions

When analyzing more than one coefficient at a time, we can use confidence intervals, or we can construct [**confidence regions**]{style="color:blue"}:

- A confidence region is the same thing as a CI, except that it may have more than one dimension.
- A confidence region provides us with a region of plausible values of our target parameter(s).  



```{r}
#library(ellipse) # required package to draw confidence region

# construct 95% joint confidence intervals for beta_area and beta_adjacent.
plot(ellipse(lmod, c(2, 6)), type="l", ylim = c(-0.13,0))

# Plot the origin and the center point
points(c(0,coef(lmod)[2]),c(0, coef(lmod)[6]), col = c("red", "blue"), pch = c(19,19))

# add vertical and horizontal lines for individual confidence intervals
abline(v = confint(lmod)[2,], lty = 2) # plots vertical lines
abline(h = confint(lmod)[6,], lty = 2) # plots horizontal lines
```


## Question 17 {#sec-q17}

Based on the confidence region above, is it plausible that $\beta_{\rm Area} = \beta_{\rm Adjacent}=0$?  Why or why not?


### Solution to Question 17


No. The red point at $(0,0)$ is not inside the ellipse, thus these values are not simultaneously plausible.

<br>  
<br>  

## Question 18 {#sec-q18}

Based on the confidence region above, is it plausible that $\beta_{\rm Area}= -0.06$ and $\beta_{\rm Adjacent}=-0.045$?  Why or why not?

### Solution to Question 18

No. The red point at $(-0.06, -0.045)$ is not inside the ellipse, thus these values are not simultaneously plausible.


## Reading Confidence Region Plots

- **Any point that lies within** the $100(1- \alpha)\%$ confidence region for $\beta_i,\beta_j, \ldots, \beta_k$ represents values of $c_i,c_j, \ldots ,c_k$ for which the associated **null hypothesis would not be rejected at significance level $\alpha$**.
- **Any point outside of the confidence region** represents values of $c_i, c_j, \ldots ,c_k$ for which the associated **null hypothesis would be rejected**.
- Both the horizontal width and vertical width of the joint confidence region is wider than the widths of the individual confidence intervals.
- The overall area of the joint region is smaller than the area of the intersection between the two individual confidence regions.
  - This is because the estimated regression parameters are positively correlated.
- If the lines of the individual confidence regions were tangential to the joint region, then the individual CIs would be jointly correct (their confidence level would be at least 95%).
- It is possible to make different conclusions when using individual confidence regions in comparison with the joint confidence regions!
 - **The joint confidence regions should be preferred**
- We must be cautious about how we interpret univariate hypothesis tests or confidence intervals because the same conclusions may not be jointly true!


# Bootstrap Confidence Intervals

The $F$ and $t$-based confidence regions and intervals we have described depend on assumptions:

- In linear regression, the errors are assumed to be normal.
- In estimating a population mean, if $n < 30$, the underlying population is assumed to be symmetric.

- In general, **parametric CI's** assume we know the sampling distribution of the statistic that estimates our target parameter. 
- How would we approximate the sampling distribution of a statistic using simulated data if we do not know the true error distribution?

**We can use the bootstrap method to produce a confidence interval for our regression coefficients when error distribution is unknown or non-normal.**


## Bootstrap Process for Komodo Dragon Weights

1. Generate a bootstrap resample (same size as original sample) by sampling with replacement from the original sample `dragon_wt`. 


```{r}
dragon_wt <- c(145, 178, 142, 139, 160, 190, 168, 122)  # observed sample
(boot_dragon <- sample(dragon_wt, size = 8, replace = TRUE)) # bootstrap resample
```

2. 	Compute the relevant statistic (mean in this case) of the bootstrap resample.

```{r}
mean(boot_dragon) 
```

3.	Repeat steps 1-2 many times (10,000 times will suffice)

```{r}
set.seed(123)
N <- 10000  # set the number of bootstrap samples to be generated
boot_dist <- numeric(N)  # place to store bootstrap statistics
n_wt <- length(dragon_wt)

for (i in 1:N){
  temp_dragon <- sample(dragon_wt, size = n_wt, replace = TRUE)  # generate bootstrap resample
  boot_dist[i] <- mean(temp_dragon)
}
```

4. Choose the the interval containing the middle 95% of the bootstrap resamples as the 95% bootstrap percentile confidence interval.

```{r}
(dragon_boot_cut <- quantile(boot_dist, probs = c(0.025, 0.975)))
```


A 95% bootstrap percentile confidence interval is from $141.125$ to $170.125$ pounds.


## Comparing with Parametric Confidence Intervals

Using parametric methods, we earlier obtains a 95% confidence interval for the mean weight of all Komodo dragons from
$145.76$ to $165.24$ pounds.

## Visualizing Bootstrap Confidence Intervals

```{r}
# plot density for bootstrap dist for sample mean weight
plot(density(boot_dist), xlab = "Mean Weight", main = "")
title("Bootstrap distribution for Mean Weight")
abline(v = c(141.125, 170.125), lty = 2)  # plot CI
```

## Bootstrap Process for Regression Coefficients

1. Generate a bootstrap resample $\boldsymbol\epsilon^{\ast}$ by sampling with replacement from $\hat{\epsilon}_1, \hat{\epsilon}_2, \ldots , \hat{\epsilon}_n$.

```{r}
resids <- residuals(lmod)  # observed residuals
as.vector(resids)
```

```{r}
n_resids <- length(resids)  # size of each sample
boot_resids <- sample(resids, size = n_resids, replace = TRUE) # bootstrap resample
as.vector(boot_resids)
```

2. Form $\mathbf{y}^{\ast} =X \boldsymbol{\hat{\beta}} + \boldsymbol\epsilon^{\ast}$ for fixed $X$ and using the $\boldsymbol{\hat{\beta}}$ from the fitted model of the original data

```{r}
new_y <- fitted(lmod) + boot_resids
```

3. 	Compute $\boldsymbol{\hat{\beta}}^{\ast}$ from $(\mathbf{X},\mathbf{y}^{\ast})$.

```{r}
boot_model <- lm(new_y ~ Area + Elevation + Nearest + Scruz + Adjacent, data = gala)
coef(boot_model)
coef(lmod)
```

4.	Repeat steps 1-3 many times (10000 times will suffice)

```{r}
set.seed(123)  # fixes random seed for consistent output
N <- 10000  # number of bootstrap samples to be generated
resids <- residuals(lmod)  # observed residuals
n_resids <- length(resids)  # number of observed residuals

# initially empty matrix with N rows and 6 columns
coefmat <- matrix(NA, N, 6)  # matrix where we will store results

preds <- fitted(lmod)  # fitted values
for (i in 1:N){
  boot_y <- preds + sample(resids, size = n_resids, replace = TRUE) # randomly assign errors
  bmod <- lm(boot_y ~ Area + Elevation + Nearest + Scruz + Adjacent, data = gala)
  coefmat[i,] <- coef(bmod)
}
```

5. Estimate the sampling distribution of the estimated coefficients using the bootstrap distribution of the estimated coefficients from the bootstrapped data sets.

```{r}
colnames(coefmat) <- c("Intercept", colnames(gala[,3:7])) # rename columns of coefmat
coefmat <- data.frame(coefmat) # convert to data frame

# construct 95% CIs for each coefficients using the apply function
cis <- apply(coefmat, 2, # 2 means apply to each column of coefmat
             quantile, probs = c(.025, .975))  
cis
```

## Comparing with Parametric Confidence Intervals


```{r}
confint(lmod)
```

## Visualizing Bootstrap Confidence Intervals


```{r}
plot(density(coefmat$Area), xlab = "Area", main = "") # plot density
title("Bootstrap distribution for betahat_Area") #title
abline(v = c(-0.06216773, 0.01836816), lty = 2) # plot CI
```

```{r}
# same thing for Adjacent
plot(density(coefmat$Adjacent), xlab = "Adjacent", main = "") # plot density
title("Bootstrap distribution for betahat_Adjacent") #title
abline(v = c(-0.10542995, -0.04009888), lty = 2) # plot CI
```

```{r}
#library(ggplot2) # same plots using ggplot2
ggplot(coefmat, aes(x = Area)) + 
  geom_density() + 
  geom_vline(xintercept = c(-0.06216773, 0.01836816), lty = 2) + 
  theme_bw()

ggplot(coefmat, aes(x = Adjacent)) + 
  geom_density() + 
  geom_vline(xintercept = c(-0.10542995, -0.04009888), lty = 2) + 
  theme_bw()
```


- Both densities are roughly symmetric and normal, though this is not always the case.
- Bootstrap methods can be used for hypothesis testing, but permutation-based methods are generally preferred.  
- There are other (more complex) methods for constructing bootstrap confidence intervals for the coefficients.


# Appendix A: Summary of Hypothesis Testing {#sec-appendA}

1.  State the  [**hypotheses**]{style="color:blue"} and identify (from the alternative claim in $H_a$) if it is a one or two-tailed test.

-   $H_0$ is the “boring” claim. Express using an equal sign $=$.
-   $H_a$ is the claim we want to show is likely true. Use inequality sign ($>$, $<$, or $\ne$).
-   State both $H_0$ and $H_a$ in terms of population parameters such as $\mu$ and $p$.

2.  Compute the  [**test statistic**]{style="color:blue"}.

-   If the observed sample contradicts the null claim, the result is significant.
-   A standardized test statistic measures how many SE's the observed stat is from the null claim.
-   A standardized test statistic with a large absolute value is supporting evidence to reject $H_0$.

3.  Using the null distribution, compute the  [**p-value**]{style="color:blue"}. The p-value is the probability of getting a sample with a test statistic as or more extreme than the observed sample assuming $H_0$ is true.

-   The p-value is the area in one or both tails beyond the test statistic.
-   The p-value is a probability, so we have $0 < \mbox{p-value} < 1$.
-   The smaller the p-value, the stronger the evidence to reject $H_0$.

4.  Based on the  [**significance level**]{style="color:blue"}, $\alpha$, make a decision to reject or not reject the null hypothesis

-   If p-value $\leq \alpha$, we reject $H_0$.
-   If p-value $> \alpha$, we do not reject $H_0$.

5. [**Summarize the results**]{style="color:blue"} in practical terms, **in the context of the example**.

-   If we reject $H_0$, this means there is enough evidence to support the claim in $H_a$.
-   If we do not reject $H_0$, this means there is not evidence to reject $H_0$ nor support $H_a$. The test is inconclusive.



