---
title: "Homework 5"
author: 
  - name: "Enter your name"
    affiliation: "University of Colorado Denver"
format: 
  html:
    embed-resources: true    
editor: source
warning: false
---

# Solutions HW 5: Due Wed, Oct 8 at 2:00 PM

## Instructions:


1.  <span style="color:red">Be sure to download and save the file so you are working with your own private copy of the file.</span>

2.  <span style="color:green">After you have copied the Quarto file onto your computer, you can begin editing and saving your work.</span>

**Submit your work electronically as two separate files:**

-   A Quarto markdown file with file extension `.qmd` with all code cells run and relevant output displayed in the document.
-   After you have completed all questions, render your completed work to an html file.
-   Submit the html file that is created with file extension `.html`.

Upload both files into the Homework 5 assignment in Canvas before <span style="color:blue">**2 PM on Wed, Oct 8.**</span>

**You must provide supporting work for your all of your answers.**

-   <span style="color:blue">That work will include using R code cells.</span>
    -   Feel free to insert as much R code as you like.
    -   When asked, you must explain how the output of a code cell is used to solve the question.
    -   Code with incomplete interpretation of the output  may not receive full credit.
    -   Code that involves methods we have not discussed this semester may not receive full credit.
-   <span style="color:blue">Your work may also involve typing explanations and justifications in text cells.</span>
    -   Be sure to type up your work/explanation so I can see how you arrived at your answer.
    -   <span style="color:blue">**When you need to type math expressions, use LaTeX.**</span>

## Important Academic Policies:


-   **Be your own work.** Though you may collaborate with others, everyone is responsible for writing up the work in their own way using their own methods. Plagiarism of any form is not tolerated.
-   **Be complete.** You must provide all work and/or explanations needed to find the solution. Answers with insufficient or incomplete supporting work may lose credit.
-   **Adhere to the Code of Academic Honesty.**
-   **Be clearly typed.** Your solution to a question must be clear and typed in complete sentences. 

### <span style="color:blue">Let me know if you need help, and GOOD LUCK!!!</span>


# Be Sure to Do the Following Once Completed

---

-   Be sure you have run ALL code cells before saving your work.
-   Be sure all relevant code output is displayed in Quarto document.
-   If you used LaTeX, make sure all of the formulas look good when rendering to html.
-   Then be sure to save the your work as a Quarto document.
-   Finally, be sure you render your final work to a static `html` that you submit along with your Quarto file in Canvas.
-   If you preferred to write work on questions 3-6, be sure to submit either a paper copy to me or upload a pdf version of your written work into Canvas along with your other documents.

# Loading Data and Required Packages


Run the code cells below to load required packages and the data frame `insur` from a csv file.

```{r}
#| message: false
#| warning: false
# load required packages 
library(tidyverse)
```

```{r}
# requires tidyverse package
insur<- read_csv("https://raw.githubusercontent.com/CU-Denver-MathStats-OER/Statistical-Theory/main/Data/insurance.csv", 
                 col_types = cols(sex = col_factor(),
                                  smoker = col_factor(),
                                  region = col_factor()))
```


## Getting to Know the Data


```{r}
glimpse(insur)
```

The data set contains information from a random sample of $1,\!338$ insurance policies. The data contains 7 variables:

- `age` of the policy holder in years.
- `sex` assigned at birth of the policy holder.
- `bmi` of the policy holder (in $\text{kg}/\text{m}^2$).
- `children` is the number of children of the policy holder.
- `smoker` is whether the policy holder is smoker or not.
- `region` is the region where the policy holder lives.
- `charges` is the total medical costs billed by the insurance company to the policy holder.




# Question 1 {#sec-q1}

Construct a side by side box plots to display the distribution of `charges` separately for each `smoker` status. Comment on any interesting features that you see.


## Solution to Question 1



<br>  
<br>  



# Question 2 {#sec-q2}

Construct a scatter plot of the data for the using `charges` as the response variable and `bmi` as the predictor variable.  Distinguish between the observations from each `smoker` status using different symbols and colors. Comment on any interesting features that you see.


## Solution to Question 2



<br>  
<br>  


# Question 3 {#sec-q3}

Fit a one-way model for `charges` using `smoker`.  Based on your model (whose coefficients you should display on screen), give the estimated mean `charges` for each `smoker` status?

## Solution to Question 3



<br>  
<br>  



# Question 4 {#sec-q4}

Interpret all of the coefficients from your model in [Question 3](#sec-q3).


## Solution to Question 4



<br>  
<br>  



# Question 5 {#sec-q5}

Fit a main-effects (parallel lines) model to `charges` using `bmi` and `smoker`.  Based on the coefficients of the output (which you should display on screen), write out separate models for each `smoker` status for insurance `charge` using `bmi` as the predictor. Provide two separate equations.


## Solution to Question 5



<br>  
<br>  



# Question 6 {#sec-q6}

Interpret all coefficients from [Question 5](#sec-q5) in practical terms in the context of this example.


## Solution to Question 6




<br>  
<br>  


# Question 7 {#sec-q7}

Fit an interaction model to `charges` using `bmi` and `smoker`.  Based on the coefficients of the output (which you should display on screen), write out separate models for each `smoker` status for insurance `charge` using `bmi` as the predictor. Provide two separate equations.

## Solution to Question 7




<br>  
<br>  


# Question 8 {#sec-q8}

Interpret all coefficients from [Question 7](#sec-q7) in practical terms in the context of this example.


## Solution to Question 8




<br>  
<br>  




# Question 9 {#sec-q9}

Create a scatter plot of `charges` vs `bmi` adding both regression lines for the **parallel lines model** you found in [Question 5](#sec-q5).


## Solution to Question 9



<br>  
<br>  


# Question 10 {#sec-q10}

Create a scatter plot of `charges` vs `bmi` adding both regression lines for the **interaction model** you found in [Question 7](#sec-q7).


## Solution to Question 10



<br>  
<br>  


# Question 11 {#sec-q11}

Which model (main effects from [Question 5](#sec-q5) or interaction from [Question 7](#sec-q7)) do you believe is more appropriate and why? Support your answer with a brief explanation. 

- Your justification can refer to previous parts of this assignment. 
- It can also refer to additional plots/analysis you want to perform that was not asked of you on this assignment. 
- You do not need to do any additional analysis, but you are welcome to do so.


## Solution to Question 11



<br>  
<br>  
