---
title: "Solutions Homework 8"
author: 
  - name: "Enter your name"
    affiliation: "University of Colorado Denver"
format: 
  html:
    embed-resources: true    
editor: source
warning: false
---

# HW 8: Due Wed, November 5 at 2:00 PM

## Instructions:


1.  <span style="color:red">Be sure to download and save the file so you are working with your own private copy of the file.</span>

2.  <span style="color:green">After you have copied the Quarto file onto your computer, you can begin editing and saving your work.</span>

**Submit your work electronically as two separate files:**

-   A Quarto markdown file with file extension `.qmd` with all code cells run and relevant output displayed in the document.
-   After you have completed all questions, render your completed work to an html file.
-   Submit the html file that is created with file extension `.html`.

Upload both files into the Homework 8 assignment in Canvas before <span style="color:blue">**2 PM on Wed, November 5.**</span>

**You must provide supporting work for your all of your answers.**

-   <span style="color:blue">That work will include using R code cells.</span>
    -   Feel free to insert as much R code as you like.
    -   When asked, you must explain how the output of a code cell is used to solve the question.
    -   Code with incomplete interpretation of the output  may not receive full credit.
    -   Code that involves methods we have not discussed this semester may not receive full credit.
-   <span style="color:blue">Your work may also involve typing explanations and justifications in text cells.</span>
    -   Be sure to type up your work/explanation so I can see how you arrived at your answer.
    -   <span style="color:blue">**When you need to type math expressions, use LaTeX.**</span>

## Important Academic Policies:


-   **Be your own work.** Though you may collaborate with others, everyone is responsible for writing up the work in their own way using their own methods. Plagiarism of any form is not tolerated.
-   **Be complete.** You must provide all work and/or explanations needed to find the solution. Answers with insufficient or incomplete supporting work may lose credit.
-   **Adhere to the Code of Academic Honesty.**
-   **Be clearly typed.** Your solution to a question must be clear and typed in complete sentences. 

### <span style="color:blue">Let me know if you need help, and GOOD LUCK!!!</span>


# Be Sure to Do the Following Once Completed

---

-   Be sure you have run ALL code cells before saving your work.
-   Be sure all relevant code output is displayed in Quarto document.
-   If you used LaTeX, make sure all of the formulas look good when rendering to html.
-   Then be sure to save the your work as a Quarto document.
-   Finally, be sure you render your final work to a static `html` that you submit along with your Quarto file in Canvas.
-   If you preferred to write work on questions 3-6, be sure to submit either a paper copy to me or upload a pdf version of your written work into Canvas along with your other documents.

# Loading Data and Required Packages

```{r}
#| message: false
#| warning: false
library(tidyverse)
```

Load the `prostate` data set from the `faraway` package. The data are from a study of 97 men with prostate cancer who were due to receive a radical prostatectomy. More details can be found by running `?faraway::prostate` in the R Console. The variables in the data set include:

- `lcavol`: log(cancer volume). Cancer volume has units $\text{cm}^3$.
- `lweight`: log(prostate weight). Prostate weight has units grams.
- `age`: Subject age (years)
- `lbph`: log(benign prostatic hyperplasia amount). BPH is graded on a scale from 0 to 35.
- `svi`: Seminal vesicle invasion indicates the extent of cancer spread (0 no spread to 1 high spread).
- `lcp`: log(capsular penetration). Capsular penetration has units of mm.
- `gleason`: Gleason score which is a value from 2 to 10.
- `pgg45`: Percentage Gleason scores 4 or 5.
- `lpsa`: log(prostate specific antigen). PSA has units ng/mL. This is the response variable.

::: {.callout-note}
As usual in statistics, we use log() to indicate the natural log (not log base 10).
:::

```{r}
data(prostate, package = "faraway")
```


We will consider the relationship between the response `lpsa` and several of the other variables in the data set.


# Question 1 

## Question 1a

Fit a linear model regressing `lpsa` on `lcavol`, `lweight`, `svi`.
Summarize the fitted model using the `summary` function.

### Solution to Question 1a


<br>  
<br>  


## Question 1b

Write the equation for the fitted model.

### Solution to Question 1b


<br>  
<br>  


# Question 2

Use the `apply` function to compute the sample means of each variable in the `prostate` data set. Provide the sample mean of each variable in `prostate` to two decimal places.

::: {.callout-tip}
See a tutorial on `apply` functions at <https://www.datacamp.com/tutorial/r-tutorial-apply-family> for help.
:::


## Solution to Question 2



<br>  
<br>  


# Question 3

Consider the linear model regressing `lpsa` on `lcavol`, `lweight`, `svi`.
Estimate the mean value of `lpsa` when `lcavol`, `lweight`, and `svi` are fixed at
the average of their observed values.


## Solution to Question 3

<br>  
<br>  



# Question 4

Consider the linear model regressing `lpsa` on `lcavol`, `lweight`, `svi`.
Construct a 90% confidence interval for the mean value of `lpsa` when `lcavol`, `lweight`, and `svi` are fixed at the average of their observed values.

## Solution to Question 4


<br>  
<br>  


# Question 5

Use the `apply` function to compute the $0.25$ quantile of each variable in the `prostate` data set. Provide the quantile of each variable in `prostate` to two decimal places.

## Solution to Question 5

<br>  
<br>  



# Question 6

Consider the linear model regressing `lpsa` on `lcavol`, `lweight`, `svi`.
Predict the `lpsa` value for a new individual with `lcavol`, `lweight`, and `svi` fixed at the $0.25$ quantiles of their observed values.

## Solution to Question 6

<br>  
<br>  


# Question 7 

Consider the linear model regressing `lpsa` on `lcavol`, `lweight`, `svi`.
Construct a 80% prediction interval for the `lpsa` value for a new individual with `lcavol`, `lweight`, and `svi` are fixed at the $0.25$ quantiles of their observed values.

## Solution to Question 7

<br>  
<br>  


# Question 8

Assuming relevant considerations are identical (same fitted model, level, regressor values, etc), which interval do you expect to be wider: a confidence interval for the mean response or a prediction interval for a new response?

## Solution to Question 8

<br>  
<br>  



# Question 9

Assume you are estimating the mean response for a fixed confidence level (say 95%). Would you expect the confidence interval to be wider when the regressor values are fixed at their sample mean or at 1% of their observed values? Why?


## Solution to Question 9

<br>  
<br>  




